---
title: "NCBI Database Retrieval"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NCBI Database Retrieval}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(width = 750)
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
```

# Retrieve Sequence Databases from NCBI

NCBI stores a variety of specialized database such as [Genbank, RefSeq, Taxonomy, SNP, etc.](https://www.ncbi.nlm.nih.gov/guide/data-software/) on their servers.
The `download.database()` and `download.database.all()` functions implemented in `biomartr` allows users to download these databases from NCBI.

## Search for available databases

Before downloading specific databases from NCBI users might want to list available databases. Using the `listNCBIDatabases()` function users can retrieve a list of  available databases stored on NCBI.

```{r, eval=TRUE}
# retrieve a list of available sequence databases at NCBI
biomartr::listNCBIDatabases(db = "all")
```

However, in case users already know which database they would like to retrieve
they can filter for the exact files by specifying the NCBI database name. In the following example all sequence files that are part of the `NCBI nr` database shall be retrieved.


First, the `listNCBIDatabases(db = "nr")` allows to list all files corresponding to the `nr` database. 

```{r, eval=TRUE}
# show all NCBI nr files
biomartr::listNCBIDatabases(db = "nr")
```


The output illustrates that the `NCBI nr` database has been separated into 41 files.

Further examples are:


```{r, eval=TRUE}
# show all NCBI nt files
biomartr::listNCBIDatabases(db = "nt")
```

```{r, eval=TRUE}
# show all NCBI ESTs others
biomartr::listNCBIDatabases(db = "est_others")
```


```{r, eval=TRUE}
# show all NCBI RefSeq (only genomes)
head(biomartr::listNCBIDatabases(db = "refseq_genomic"), 20)
```


```{r, eval=TRUE}
# show all NCBI RefSeq (only proteomes)
biomartr::listNCBIDatabases(db = "refseq_protein")
```

```{r, eval=TRUE}
# show all NCBI RefSeq (only RNA)
biomartr::listNCBIDatabases(db = "refseq_rna")
```

```{r, eval=TRUE}
# show NCBI swissprot
biomartr::listNCBIDatabases(db = "swissprot")
```

```{r, eval=TRUE}
# show NCBI PDB
biomartr::listNCBIDatabases(db = "pdb")
```

```{r, eval=TRUE}
# show NCBI Human database
biomartr::listNCBIDatabases(db = "human")
```

```{r, eval=TRUE}
# show NCBI EST databases
biomartr::listNCBIDatabases(db = "est")
```

__Please not that all lookup and retrieval function will only work properly when a sufficient internet connection is provided.__

In a next step users can use the `listNCBIDatabases()` and `download.database.all()` functions to retrieve 
all files corresponding to a specific NCBI database.

## Download available databases

Using the same search strategy by specifying the database name as described above, users can now download these databases using the `download.database()` function.

For downloading only single files users can type:

```{r, eval=FALSE}
# select NCBI nr version 27 =  "nr.27.tar.gz"
# and download it into the folder
biomartr::download.database(db = "nr.27.tar.gz", path = "nr")
```

Using this command first a folder named `nr` is created and the
file `nr.27.tar.gz` is downloaded to this folder. 
This command will download the pre-formatted (by makeblastdb formatted) database version is retrieved.

Alternatively, users can retrieve all `nr` files with one command by typing:

```{r, eval=FALSE}
# download the entire NCBI nr database
biomartr::download.database.all(db = "nr", path = "nr")
```

Using this command, all `NCBI nr` files are loaded into the `nr` folder (`path = "nr"`).


The same approach can be applied to all other databases mentioned above, e.g.:

```{r, eval=FALSE}
# download the entire NCBI nt database
biomartr::download.database.all(db = "nt", path = "nt")
```

```{r, eval=FALSE}
# download the entire NCBI refseq (protein) database
biomartr::download.database.all(db = "refseq_protein", path = "refseq_protein")
```

```{r, eval=FALSE}
# download the entire NCBI PDB database
biomartr::download.database.all(db = "pdb", path = "pdb")
```

__Please notice that most of these databases are very large, so users should take of of providing a stable internect connection throughout the download process.__
