---
title: "Functional Annotation with biomartr"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functional Annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(width = 750)
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
```


## Functional Annotation Retrieval from `Ensembl Biomart`

The [Ensembl Biomart](http://ensemblgenomes.org/info/access/biomart) database enables users to retrieve a vast diversity of annotation data
for specific organisms. Initially, Steffen Durinck and Wolfgang Huber provided a powerful interface between
the R language and [Ensembl Biomart](http://ensemblgenomes.org/info/access/biomart) by implementing the R package [biomaRt](http://www.bioconductor.org/packages/release/bioc/html/biomaRt.html). 

The purpose of the `biomaRt` package was to mimic the ENSEMBL BioMart database structure to construct queries that can be sent to the Application Programming Interface (API) of BioMart. Although, this procedure was very useful in the past, it seems not intuitive from an organism centric point of view. Usually, users wish to download functional annotation for a particular organism of interest. However, the BioMart and thus the `biomaRt` package require that users already know in which `mart` and  `dataset` the organism of interest will be found which requires significant efforts of searching and screening. In addition, once the  `mart` and  `dataset` of a particular organism of interest were found and specified the user must again learn which `attribute` has to be specified to retrieve the functional annotation information of interest. 

The new functionality implemented in the `biomartr` package aims to overcome this 
search bottleneck by extending the functionality of the [biomaRt](http://www.bioconductor.org/packages/release/bioc/html/biomaRt.html) package. The new `biomartr` package introduces a more organism cantered annotation retrieval concept which does not require to screen for `marts`, `datasets`, and `attributes` beforehand. With `biomartr` users only need to specify the `scientific name` of the organism of interest to then retrieve available `marts`, `datasets`, and `attributes` for the corresponding organism of interest.   

This paradigm shift enables users to quickly construct queries to the BioMart database without having to learn the particular database structure and organization of BioMart. 

The following sections will introduce users to the functionality and data retrieval precedures of `biomartr` and will show how `biomartr`
extends the functionality of the initial [biomaRt](http://www.bioconductor.org/packages/release/bioc/html/biomaRt.html) package.

### The old `biomaRt` query methodology		

The best way to get started with the _old_ methodology presented by the established [biomaRt](http://www.bioconductor.org/packages/release/bioc/html/biomaRt.html) package is to understand the workflow of its data retrieval process. The query logic of the `biomaRt` package derives from the database organization of [Ensembl Biomart](http://ensemblgenomes.org/info/access/biomart) which stores a vast diversity of annotation data
for specific organisms. In detail, the [Ensembl Biomart](http://ensemblgenomes.org/info/access/biomart) database is organized into so called:		
 `marts`, `datasets`, and `attributes`. `Marts` denote a higher level category of functional annotation such as `SNP` (e.g. for functional annotation of particular single nucleotide polymorphisms (SNPs)) or `FUNCGEN` (e.g. for functional annotation of regulatory regions or relationsships of genes). `Datasets` denote the
particular species of interest for which functional annotation is available __within__ this specific `mart`. It can happen that
`datasets` (= particular species of interest) are available in one `mart` (= higher category of functional annotation) but not in an other `mart`.
For the actual retrieval of functional annotation information users must then specify the `type` of functional annotation information they 
wish to retrieve. These `types` are called `attributes` in the `biomaRt` notation.
 
Hence, when users wish to retrieve information for a specific organism of interest, they first need to specify a particular `mart` and `dataset` in which the information of the corresponding organism of interest can be found. Subsequently they can specify the `attributes` argument to retrieve a particular type of functional annotation (e.g. Gene Ontology terms).

The following section shall illustrate how `marts`, `datasets`, and `attributes` could be explored
using `biomaRt` before the `biomartr` package existed.
 		
The availability of `marts`, `datasets`, and `attributes` can be checked by the following functions:		

```{r,eval=TRUE}		
# install the biomaRt package		
# source("http://bioconductor.org/biocLite.R")		
# biocLite("biomaRt")		
# load biomaRt		
library(biomaRt)		
# look at top 10 databases		
head(biomaRt::listMarts(host = "www.ensembl.org"), 10)		
```	

Users will observe that several `marts` providing annotation for specific classes of organisms or groups of organisms are available.		

For our example, we will choose the `hsapiens_gene_ensembl` `mart` and list all available datasets that are element of this `mart`.		

```{r,eval=TRUE}		
head(biomaRt::listDatasets(biomaRt::useMart("ENSEMBL_MART_ENSEMBL", host = "www.ensembl.org")), 10)		
```		

The `useMart()` function is a wrapper function provided by `biomaRt` to connect a selected BioMart database (`mart`) with a corresponding dataset stored within this `mart`.		
We select dataset `hsapiens_gene_ensembl` and now check for available attributes (annotation data) that can be accessed for `Homo sapiens` genes.	

```{r,eval=FALSE}		
head(biomaRt::listAttributes(biomaRt::useDataset(
                                         dataset = "hsapiens_gene_ensembl", 		
                                         mart    = useMart("ENSEMBL_MART_ENSEMBL",		
                                         host    = "www.ensembl.org"))), 10)		
```		
		
Please note the nested structure of this attribute query. For an attribute query procedure an additional wrapper function named `useDataset()` is needed in which `useMart()` and a corresponding dataset needs to be specified. The result is a table storing the name of available attributes for		
_Homo sapiens_ as well as a short description.		

Furthermore, users can retrieve all filters for _Homo sapiens_ that can be specified by the actual BioMart query process.		
		
```{r,eval=FALSE}		
 head(biomaRt::listFilters(biomaRt::useDataset(dataset = "hsapiens_gene_ensembl", 		
                                               mart    = useMart("ENSEMBL_MART_ENSEMBL",		
                                               host    = "www.ensembl.org"))), 10)		
```		
 		
After accumulating all this information, it is now possible to perform an actual BioMart query by using the `getBM()` function.		
 		
In this example we will retrieve attributes: `start_position`,`end_position` and `description`		
for the _Homo sapiens_ gene `"GUCA2A"`.		
 		
Since the input genes are `ensembl gene ids`, we need to specify the `filters` argument `filters = "hgnc_symbol"`.		
 		
```{r,eval=FALSE}		
 # 1) select a mart and data set		
 mart <- biomaRt::useDataset(dataset = "hsapiens_gene_ensembl", 		
                    mart    = useMart("ENSEMBL_MART_ENSEMBL",		
                    host    = "www.ensembl.org"))		
 		
 # 2) run a biomart query using the getBM() function		
 # and specify the attributes and filter arguments		
 geneSet <- "GUCA2A"		
 		
 resultTable <- biomaRt::getBM(attributes = c("start_position","end_position","description"),		
                      filters    = "hgnc_symbol", 		
                      values     = geneSet, 		
                      mart       = mart)		
 		
 resultTable 		
```		
 		
When using `getBM()` users can pass all attributes retrieved by `listAttributes()` to the `attributes` argument of the `getBM()` function.		
  
  
## Extending `biomaRt` using the new query system of the `biomartr` package

### Getting Started with `biomartr`

This query methodology provided by `Ensembl Biomart` and the `biomaRt` package is a very well defined approach
for accurate annotation retrieval. Nevertheless, when learning this query methodology it (subjectively)
seems non-intuitive from the user perspective. Therefore, the `biomartr` package provides another
query methodology that aims to be more organism centric.

Taken together, the following workflow allows users to perform fast BioMart queries for 
attributes using the `biomart()` function implemented in this `biomartr` package:

1) get attributes, datasets, and marts via : `organismAttributes()`

2) choose available biological features (filters) via: `organismFilters()`

3) specify a set of query genes: e.g. retrieved with `getGenome()`, `getProteome()` or `getCDS()`

4) specify all arguments of the `biomart()` function using steps 1) - 3) and
perform a BioMart query

__Note that dataset names change very frequently due to the update of dataset versions.
So in case some query functions do not work properly, users should check with
`organismAttributes(update = TRUE)` whether or not their dataset name has been changed.
For example, `organismAttributes("Homo sapiens", topic = "id", update = TRUE)`
might reveal that the dataset `ENSEMBL_MART_ENSEMBL` has changed.__


## Retrieve marts, datasets, attributes, and filters with biomartr

### Retrieve Available Marts

The `getMarts()` function allows users to list all available databases that can be accessed through BioMart interfaces.

```{r,eval=FALSE}
# load the biomartr package
library(biomartr)

# list all available databases
biomartr::getMarts()
```

```
  mart                  version                       
   <chr>                 <chr>                         
 1 ENSEMBL_MART_ENSEMBL  Ensembl Genes 92              
 2 ENSEMBL_MART_MOUSE    Mouse strains 92              
 3 ENSEMBL_MART_SEQUENCE Sequence                      
 4 ENSEMBL_MART_ONTOLOGY Ontology                      
 5 ENSEMBL_MART_GENOMIC  Genomic features 92           
 6 ENSEMBL_MART_SNP      Ensembl Variation 92          
 7 ENSEMBL_MART_FUNCGEN  Ensembl Regulation 92         
 8 plants_mart           Ensembl Plants Genes 39       
 9 plants_variations     Ensembl Plants Variations 39  
10 fungi_mart            Ensembl Fungi Genes 39        
11 fungi_variations      Ensembl Fungi Variations 39   
12 protists_mart         Ensembl Protists Genes 39     
13 protists_variations   Ensembl Protists Variations 39
14 metazoa_mart          Ensembl Metazoa Genes 39      
15 metazoa_variations    Ensembl Metazoa Variations 39 
```

### Retrieve Available Datasets from a Specific Mart

Now users can select a specific database to list all available datasets
that can be accessed through this database. In this example we choose
the `ENSEMBL_MART_ENSEMBL` database.

```{r,eval=FALSE}
head(biomartr::getDatasets(mart = "ENSEMBL_MART_ENSEMBL") , 5)
```

```
 dataset                description                         version        
  <chr>                  <chr>                               <chr>          
1 pabelii_gene_ensembl   Orangutan genes (PPYG2)             PPYG2          
2 mpahari_gene_ensembl   Shrew mouse genes (PAHARI_EIJ_v1.1) PAHARI_EIJ_v1.1
3 mspretus_gene_ensembl  Algerian mouse genes (SPRET_EiJ_v1) SPRET_EiJ_v1   
4 chircus_gene_ensembl   Goat genes (ARS1)                   ARS1           
5 trubripes_gene_ensembl Fugu genes (FUGU 4.0)               FUGU 4.0  
```

Now you can select the dataset `hsapiens_gene_ensembl` and list all available attributes that can be retrieved from this dataset.

```{r,eval=FALSE}
tail(biomartr::getDatasets(mart = "ENSEMBL_MART_ENSEMBL") , 38)
```

```
1 xtropicalis_gene_ensembl Xenopus genes (JGI 4.2)                JGI 4.2     
 2 saraneus_gene_ensembl    Shrew genes (sorAra1)                  sorAra1     
 3 nleucogenys_gene_ensembl Gibbon genes (Nleu_3.0)                Nleu_3.0    
 4 vpacos_gene_ensembl      Alpaca genes (vicPac1)                 vicPac1     
 5 pcoquereli_gene_ensembl  Coquerel's sifaka genes (Pcoq_1.0)     Pcoq_1.0    
 6 ggallus_gene_ensembl     Chicken genes (Gallus_gallus-5.0)      Gallus_gall…
 7 csavignyi_gene_ensembl   C.savignyi genes (CSAV 2.0)            CSAV 2.0    
 8 ngalili_gene_ensembl     Upper Galilee mountains blind mole ra… S.galili_v1…
 9 rroxellana_gene_ensembl  Golden snub-nosed monkey genes (Rrox_… Rrox_v1     
10 mnemestrina_gene_ensembl Pig-tailed macaque genes (Mnem_1.0)    Mnem_1.0    
# ... with 28 more rows
```

### Retrieve Available Attributes from a Specific Dataset

Now that you have selected a database (`hsapiens_gene_ensembl`) and a dataset (`hsapiens_gene_ensembl`),
users can list all available attributes for this dataset using the `getAttributes()` function.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# list all available attributes for dataset: hsapiens_gene_ensembl
head( biomartr::getAttributes(mart    = "ENSEMBL_MART_ENSEMBL", 
                              dataset = "hsapiens_gene_ensembl"), 10 )
```

```
                            name                  description
1                ensembl_gene_id               Gene stable ID
2        ensembl_gene_id_version       Gene stable ID version
3          ensembl_transcript_id         Transcript stable ID
4  ensembl_transcript_id_version Transcript stable ID version
5             ensembl_peptide_id            Protein stable ID
6     ensembl_peptide_id_version    Protein stable ID version
7                ensembl_exon_id               Exon stable ID
8                    description             Gene description
9                chromosome_name     Chromosome/scaffold name
10                start_position              Gene start (bp)
```

### Retrieve Available Filters from a Specific Dataset

Finally, the `getFilters()` function allows users to list available filters
for a specific dataset that can be used for a `biomart()` query.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# list all available filters for dataset: hsapiens_gene_ensembl
head( biomartr::getFilters(mart    = "ENSEMBL_MART_ENSEMBL", 
                           dataset = "hsapiens_gene_ensembl"), 10 )
```

```
                 name                            description
1     chromosome_name               Chromosome/scaffold name
2               start                                  Start
3                 end                                    End
4          band_start                             Band Start
5            band_end                               Band End
6        marker_start                           Marker Start
7          marker_end                             Marker End
8       encode_region                          Encode region
9              strand                                 Strand
10 chromosomal_region e.g. 1:100:10000:-1, 1:100000:200000:1
```

## Organism Specific Retrieval of Information

In most use cases, users will work with a single or a set of model organisms. In this process they will mostly be
interested in specific annotations for this particular model organism. The `organismBM()`
function addresses this issue and provides users with an organism centric query to `marts` and `datasets`
which are available for a particular organism of interest.


__Note__ that when running the following functions for the first time, the data retrieval procedure will take some time, due to the remote access to BioMart. The corresponding result is then saved in a `*.txt` file named `_biomart/listDatasets.txt` within the `tempdir()` folder, allowing subsequent queries to be performed much faster.
The `tempdir()` folder, however, will be deleted after a new R session was established. In this case
the inital call of the subsequent functions again will take time to retrieve all organism specific data from the BioMart database.

This concept of locally storing all organism specific database linking information available in BioMart into
an internal file allows users to significantly speed up subsequent retrieval queries for that particular organism.


```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# retrieving all available datasets and biomart connections for
# a specific query organism (scientific name)
biomartr::organismBM(organism = "Homo sapiens")
```

```
   organism_name description                    mart      dataset       version
   <chr>         <chr>                          <chr>     <chr>         <chr>  
 1 hsapiens      Human genes (GRCh38.p12)       ENSEMBL_… hsapiens_gen… GRCh38…
 2 hsapiens      Human sequences (GRCh38.p12)   ENSEMBL_… hsapiens_gen… GRCh38…
 3 hsapiens      marker_feature                 ENSEMBL_… hsapiens_mar… GRCh38…
 4 hsapiens      karyotype_start                ENSEMBL_… hsapiens_kar… GRCh38…
 5 hsapiens      encode                         ENSEMBL_… hsapiens_enc… GRCh38…
 6 hsapiens      marker_feature_end             ENSEMBL_… hsapiens_mar… GRCh38…
 7 hsapiens      karyotype_end                  ENSEMBL_… hsapiens_kar… GRCh38…
 8 hsapiens      Human Short Variants (SNPs an… ENSEMBL_… hsapiens_snp  GRCh38…
 9 hsapiens      Human Somatic Short Variants … ENSEMBL_… hsapiens_snp… GRCh38…
10 hsapiens      Human Somatic Structural Vari… ENSEMBL_… hsapiens_str… GRCh38…
11 hsapiens      Human Structural Variants (GR… ENSEMBL_… hsapiens_str… GRCh38…
12 hsapiens      Human Regulatory Features (GR… ENSEMBL_… hsapiens_reg… GRCh38…
13 hsapiens      Human miRNA Target Regions (G… ENSEMBL_… hsapiens_mir… GRCh38…
14 hsapiens      Human Binding Motifs (GRCh38.… ENSEMBL_… hsapiens_mot… GRCh38…
15 hsapiens      Human Other Regulatory Region… ENSEMBL_… hsapiens_ext… GRCh38…
16 hsapiens      Human Regulatory Evidence (GR… ENSEMBL_… hsapiens_peak GRCh38…
```

The result is a table storing all `marts` and `datasets` from which annotations can be retrieved
for _Homo sapiens_. Furthermore, a short description as well as the version of the dataset
being accessed (very useful for publications) is returned.

Users will observe that 3 different `marts` provide 6 different `datasets` storing annotation information for
_Homo sapiens_.

__Please note, however, that scientific names of organisms must be written correctly! For ex. "Homo Sapiens" will be treated differently (not recognized) than "Homo sapiens" (recognized).__

Similar to the `biomaRt` package query methodology, users need to specify `attributes` and `filters` to be able to perform
accurate BioMart queries. Here the functions `organismAttributes()` and `organismFilters()` provide useful and intuitive
concepts to obtain this information.


```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# return available attributes for "Homo sapiens"
head(biomartr::organismAttributes("Homo sapiens"), 20)
```

```
1 ensembl_gene_id               Gene stable ID         hsapiens_ge… ENSEMBL_M…
 2 ensembl_gene_id_version       Gene stable ID version hsapiens_ge… ENSEMBL_M…
 3 ensembl_transcript_id         Transcript stable ID   hsapiens_ge… ENSEMBL_M…
 4 ensembl_transcript_id_version Transcript stable ID … hsapiens_ge… ENSEMBL_M…
 5 ensembl_peptide_id            Protein stable ID      hsapiens_ge… ENSEMBL_M…
 6 ensembl_peptide_id_version    Protein stable ID ver… hsapiens_ge… ENSEMBL_M…
 7 ensembl_exon_id               Exon stable ID         hsapiens_ge… ENSEMBL_M…
 8 description                   Gene description       hsapiens_ge… ENSEMBL_M…
 9 chromosome_name               Chromosome/scaffold n… hsapiens_ge… ENSEMBL_M…
10 start_position                Gene start (bp)        hsapiens_ge… ENSEMBL_M…
11 end_position                  Gene end (bp)          hsapiens_ge… ENSEMBL_M…
12 strand                        Strand                 hsapiens_ge… ENSEMBL_M…
13 band                          Karyotype band         hsapiens_ge… ENSEMBL_M…
14 transcript_start              Transcript start (bp)  hsapiens_ge… ENSEMBL_M…
15 transcript_end                Transcript end (bp)    hsapiens_ge… ENSEMBL_M…
16 transcription_start_site      Transcription start s… hsapiens_ge… ENSEMBL_M…
17 transcript_length             Transcript length (in… hsapiens_ge… ENSEMBL_M…
18 transcript_tsl                Transcript support le… hsapiens_ge… ENSEMBL_M…
19 transcript_gencode_basic      GENCODE basic annotat… hsapiens_ge… ENSEMBL_M…
20 transcript_appris             APPRIS annotation      hsapiens_ge… ENSEMBL_M…
```

Users will observe that the `organismAttributes()` function returns a data.frame storing attribute names, datasets, and marts which
are available for `Homo sapiens`. After the ENSEMBL release 87 the `ENSEMBL_MART_SEQUENCE` service provided
by Ensembl does not work properly and thus the `organismAttributes()` function prints out warning messages to make the user aware
when certain marts provided bt Ensembl do not work properly, yet. 

An additional feature provided by `organismAttributes()` is the `topic` argument. The `topic` argument allows users to to search for specific attributes,  topics, or categories for faster filtering.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for attribute topic "id"
head(biomartr::organismAttributes("Homo sapiens", topic = "id"), 20)
```

```
  name                          description            dataset      mart      
   <chr>                         <chr>                  <chr>        <chr>     
 1 ensembl_gene_id               Gene stable ID         hsapiens_ge… ENSEMBL_M…
 2 ensembl_gene_id_version       Gene stable ID version hsapiens_ge… ENSEMBL_M…
 3 ensembl_transcript_id         Transcript stable ID   hsapiens_ge… ENSEMBL_M…
 4 ensembl_transcript_id_version Transcript stable ID … hsapiens_ge… ENSEMBL_M…
 5 ensembl_peptide_id            Protein stable ID      hsapiens_ge… ENSEMBL_M…
 6 ensembl_peptide_id_version    Protein stable ID ver… hsapiens_ge… ENSEMBL_M…
 7 ensembl_exon_id               Exon stable ID         hsapiens_ge… ENSEMBL_M…
 8 study_external_id             Study external refere… hsapiens_ge… ENSEMBL_M…
 9 go_id                         GO term accession      hsapiens_ge… ENSEMBL_M…
10 dbass3_id                     DataBase of Aberrant … hsapiens_ge… ENSEMBL_M…
11 dbass5_id                     DataBase of Aberrant … hsapiens_ge… ENSEMBL_M…
12 hgnc_id                       HGNC ID                hsapiens_ge… ENSEMBL_M…
13 protein_id                    INSDC protein ID       hsapiens_ge… ENSEMBL_M…
14 mim_morbid_description        MIM morbid description hsapiens_ge… ENSEMBL_M…
15 mim_morbid_accession          MIM morbid accession   hsapiens_ge… ENSEMBL_M…
16 mirbase_id                    miRBase ID             hsapiens_ge… ENSEMBL_M…
17 refseq_peptide                RefSeq peptide ID      hsapiens_ge… ENSEMBL_M…
18 refseq_peptide_predicted      RefSeq peptide predic… hsapiens_ge… ENSEMBL_M…
19 wikigene_id                   WikiGene ID            hsapiens_ge… ENSEMBL_M…
20 mobidblite                    MobiDBLite             hsapiens_ge… ENSEMBL_M…
```

Now, all `attribute names` having `id` as part of their `name` are being returned.

Another example is `topic = "homolog"`.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for attribute topic "homolog"
head(biomartr::organismAttributes("Homo sapiens", topic = "homolog"), 20)
```

```
  <chr>                                         <chr>           <chr>   <chr> 
 1 mspretus_homolog_ensembl_gene                 Algerian mouse… hsapie… ENSEM…
 2 mspretus_homolog_associated_gene_name         Algerian mouse… hsapie… ENSEM…
 3 mspretus_homolog_ensembl_peptide              Algerian mouse… hsapie… ENSEM…
 4 mspretus_homolog_chromosome                   Algerian mouse… hsapie… ENSEM…
 5 mspretus_homolog_chrom_start                  Algerian mouse… hsapie… ENSEM…
 6 mspretus_homolog_chrom_end                    Algerian mouse… hsapie… ENSEM…
 7 mspretus_homolog_canonical_transcript_protein Query protein … hsapie… ENSEM…
 8 mspretus_homolog_subtype                      Last common an… hsapie… ENSEM…
 9 mspretus_homolog_orthology_type               Algerian mouse… hsapie… ENSEM…
10 mspretus_homolog_perc_id                      %id. target Al… hsapie… ENSEM…
11 mspretus_homolog_perc_id_r1                   %id. query gen… hsapie… ENSEM…
12 mspretus_homolog_goc_score                    Algerian mouse… hsapie… ENSEM…
13 mspretus_homolog_wga_coverage                 Algerian mouse… hsapie… ENSEM…
14 mspretus_homolog_dn                           dN with Algeri… hsapie… ENSEM…
15 mspretus_homolog_ds                           dS with Algeri… hsapie… ENSEM…
16 mspretus_homolog_orthology_confidence         Algerian mouse… hsapie… ENSEM…
17 vpacos_homolog_ensembl_gene                   Alpaca gene st… hsapie… ENSEM…
18 vpacos_homolog_associated_gene_name           Alpaca gene na… hsapie… ENSEM…
19 vpacos_homolog_ensembl_peptide                Alpaca protein… hsapie… ENSEM…
20 vpacos_homolog_chromosome                     Alpaca chromos… hsapie… ENSEM…
```

Or `topic = "dn"` and `topic = "ds"` for `dn` and `ds` value retrieval.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for attribute topic "dn"
head(biomartr::organismAttributes("Homo sapiens", topic = "dn"))
```

```
  name                  description            dataset               mart      
  <chr>                 <chr>                  <chr>                 <chr>     
1 cdna_coding_start     cDNA coding start      hsapiens_gene_ensembl ENSEMBL_M…
2 cdna_coding_end       cDNA coding end        hsapiens_gene_ensembl ENSEMBL_M…
3 mspretus_homolog_dn   dN with Algerian mouse hsapiens_gene_ensembl ENSEMBL_M…
4 vpacos_homolog_dn     dN with Alpaca         hsapiens_gene_ensembl ENSEMBL_M…
5 pformosa_homolog_dn   dN with Amazon molly   hsapiens_gene_ensembl ENSEMBL_M…
6 cpalliatus_homolog_dn dN with Angola colobus hsapiens_gene_ensembl ENSEMBL_M…
```

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for attribute topic "ds"
head(biomartr::organismAttributes("Homo sapiens", topic = "ds"))
```

```
  name                description            dataset               mart        
  <chr>               <chr>                  <chr>                 <chr>       
1 ccds                CCDS ID                hsapiens_gene_ensembl ENSEMBL_MAR…
2 cds_length          CDS Length             hsapiens_gene_ensembl ENSEMBL_MAR…
3 cds_start           CDS start              hsapiens_gene_ensembl ENSEMBL_MAR…
4 cds_end             CDS end                hsapiens_gene_ensembl ENSEMBL_MAR…
5 mspretus_homolog_ds dS with Algerian mouse hsapiens_gene_ensembl ENSEMBL_MAR…
6 vpacos_homolog_ds   dS with Alpaca         hsapiens_gene_ensembl ENSEMBL_MAR…
```

Analogous to the `organismAttributes()` function, the `organismFilters()` function returns
all filters that are available for a query organism of interest.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# return available filters for "Homo sapiens"
head(biomartr::organismFilters("Homo sapiens"), 20)
```

```
   name                                description          dataset    mart    
   <chr>                               <chr>                <chr>      <chr>   
 1 chromosome_name                     Chromosome/scaffold… hsapiens_… ENSEMBL…
 2 start                               Start                hsapiens_… ENSEMBL…
 3 end                                 End                  hsapiens_… ENSEMBL…
 4 band_start                          Band Start           hsapiens_… ENSEMBL…
 5 band_end                            Band End             hsapiens_… ENSEMBL…
 6 marker_start                        Marker Start         hsapiens_… ENSEMBL…
 7 marker_end                          Marker End           hsapiens_… ENSEMBL…
 8 encode_region                       Encode region        hsapiens_… ENSEMBL…
 9 strand                              Strand               hsapiens_… ENSEMBL…
10 chromosomal_region                  e.g. 1:100:10000:-1… hsapiens_… ENSEMBL…
11 with_ccds                           With CCDS ID(s)      hsapiens_… ENSEMBL…
12 with_chembl                         With ChEMBL ID(s)    hsapiens_… ENSEMBL…
13 with_clone_based_ensembl_gene       With Clone-based (E… hsapiens_… ENSEMBL…
14 with_clone_based_ensembl_transcript With Clone-based (E… hsapiens_… ENSEMBL…
15 with_dbass3                         With DataBase of Ab… hsapiens_… ENSEMBL…
16 with_dbass5                         With DataBase of Ab… hsapiens_… ENSEMBL…
17 with_ens_hs_transcript              With Ensembl Human … hsapiens_… ENSEMBL…
18 with_ens_hs_translation             With Ensembl Human … hsapiens_… ENSEMBL…
19 with_entrezgene_trans_name          With EntrezGene tra… hsapiens_… ENSEMBL…
20 with_embl                           With European Nucle… hsapiens_… ENSEMBL…
```

The `organismFilters()` function also allows users to search for filters that correspond to
a specific topic or category.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for filter topic "id"
head(biomartr::organismFilters("Homo sapiens", topic = "id"), 20)
```

```
   name                          description               dataset     mart    
   <chr>                         <chr>                     <chr>       <chr>   
 1 with_protein_id               With INSDC protein ID ID… hsapiens_g… ENSEMBL…
 2 with_mim_morbid               With MIM morbid ID(s)     hsapiens_g… ENSEMBL…
 3 with_refseq_peptide           With RefSeq peptide ID(s) hsapiens_g… ENSEMBL…
 4 with_refseq_peptide_predicted With RefSeq peptide pred… hsapiens_g… ENSEMBL…
 5 ensembl_gene_id               Gene stable ID(s) [e.g. … hsapiens_g… ENSEMBL…
 6 ensembl_gene_id_version       Gene stable ID(s) with v… hsapiens_g… ENSEMBL…
 7 ensembl_transcript_id         Transcript stable ID(s) … hsapiens_g… ENSEMBL…
 8 ensembl_transcript_id_version Transcript stable ID(s) … hsapiens_g… ENSEMBL…
 9 ensembl_peptide_id            Protein stable ID(s) [e.… hsapiens_g… ENSEMBL…
10 ensembl_peptide_id_version    Protein stable ID(s) wit… hsapiens_g… ENSEMBL…
11 ensembl_exon_id               Exon ID(s) [e.g. ENSE000… hsapiens_g… ENSEMBL…
12 dbass3_id                     DataBase of Aberrant 3' … hsapiens_g… ENSEMBL…
13 dbass5_id                     DataBase of Aberrant 5' … hsapiens_g… ENSEMBL…
14 hgnc_id                       HGNC ID(s) [e.g. HGNC:10… hsapiens_g… ENSEMBL…
15 protein_id                    INSDC protein ID(s) [e.g… hsapiens_g… ENSEMBL…
16 mim_morbid_accession          MIM morbid accession(s) … hsapiens_g… ENSEMBL…
17 mirbase_id                    miRBase ID(s) [e.g. hsa-… hsapiens_g… ENSEMBL…
18 refseq_peptide                RefSeq peptide ID(s) [e.… hsapiens_g… ENSEMBL…
19 refseq_peptide_predicted      RefSeq peptide predicted… hsapiens_g… ENSEMBL…
20 wikigene_id                   WikiGene ID(s) [e.g. 1]   hsapiens_g… ENSEMBL…
```

## Construct BioMart queries with `biomartr`

The short introduction to the functionality of
`organismBM()`, `organismAttributes()`, and `organismFilters()`
will allow users to perform BioMart queries in a very intuitive 
organism centric way. The main function to perform BioMart queries
is `biomart()`.


For the following examples we will assume that we are interested in the annotation of specific genes from the _Homo sapiens_ proteome. We want to map the corresponding refseq gene id to a set of other gene ids used in other databases. For this purpose, first we need consult the `organismAttributes()` function.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)

head(biomartr::organismAttributes("Homo sapiens", topic = "id"))
```
```
 name                          description                  dataset    mart   
  <chr>                         <chr>                        <chr>      <chr>  
1 ensembl_gene_id               Gene stable ID               hsapiens_… ENSEMB…
2 ensembl_gene_id_version       Gene stable ID version       hsapiens_… ENSEMB…
3 ensembl_transcript_id         Transcript stable ID         hsapiens_… ENSEMB…
4 ensembl_transcript_id_version Transcript stable ID version hsapiens_… ENSEMB…
5 ensembl_peptide_id            Protein stable ID            hsapiens_… ENSEMB…
6 ensembl_peptide_id_version    Protein stable ID version    hsapiens_… ENSEMB…
```

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# retrieve the proteome of Homo sapiens from refseq
file_path <- biomartr::getProteome( db       = "refseq",
                                    organism = "Homo sapiens",
                                    path     = file.path("_ncbi_downloads","proteomes") )

Hsapiens_proteome <- biomartr::read_proteome(file_path, format = "fasta")

# remove splice variants from id
gene_set <- unlist(sapply(strsplit(Hsapiens_proteome@ranges@NAMES[1:5], ".",fixed = TRUE), function(x) x[1]))

result_BM <- biomartr::biomart( genes      = gene_set, # genes were retrieved using biomartr::getGenome()
                                mart       = "ENSEMBL_MART_ENSEMBL", # marts were selected with biomartr::getMarts()
                                dataset    = "hsapiens_gene_ensembl", # datasets were selected with biomartr::getDatasets()
                                attributes = c("ensembl_gene_id","ensembl_peptide_id"), # attributes were selected with biomartr::getAttributes()
                                filters    = "refseq_peptide") # specify what ID type was stored in the fasta file retrieved with biomartr::getGenome()

result_BM 
```

```
  refseq_peptide ensembl_gene_id ensembl_peptide_id
1      NP_000005 ENSG00000175899    ENSP00000323929
2      NP_000006 ENSG00000156006    ENSP00000286479
3      NP_000007 ENSG00000117054    ENSP00000359878
4      NP_000008 ENSG00000122971    ENSP00000242592
5      NP_000009 ENSG00000072778    ENSP00000349297
```


The `biomart()` function takes as arguments a set of genes (gene ids specified in the `filter` argument), the corresponding `mart` and `dataset`, as well as the `attributes` which shall be returned.

## Gene Ontology 

The `biomartr` package also enables a fast and intuitive retrieval of GO terms
and additional information via the `getGO()` function. Several databases can be selected
to retrieve GO annotation information for a set of query genes. So far, the `getGO()`
function allows GO information retrieval from the [Ensembl Biomart](http://ensemblgenomes.org/info/access/biomart) database. 

In this example we will retrieve GO information for a set of _Homo sapiens_ genes
stored as `hgnc_symbol`.

### GO Annotation Retrieval via BioMart

The `getGO()` function takes several arguments as input to retrieve GO information from BioMart. 
First, the scientific name of the `organism` of interest needs to be specified. Furthermore, a set of
`gene ids` as well as their corresponding `filter` notation (`GUCA2A` gene ids have `filter` notation `hgnc_symbol`; see `organismFilters()` for details)
need to be specified. The `database` argument then defines the database from which GO information shall be retrieved.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for GO terms of an example Homo sapiens gene
GO_tbl <- biomartr::getGO(organism = "Homo sapiens", 
                          genes    = "GUCA2A",
                          filters  = "hgnc_symbol")

GO_tbl
```


Hence, for each _gene id_ the resulting table stores all annotated GO terms found in [Ensembl Biomart](http://ensemblgenomes.org/info/access/biomart).

