---
title: "Functional Annotation with Ensembl Biomart, GO, and KeGG"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functional Annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(width = 750)
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
```


## Functional Annotation with `Ensembl Biomart`

The [Ensembl Biomart](http://ensemblgenomes.org/info/access/biomart) database enables users to retrieve a vast diversity of annotation data
for specific organisms. Initially, Steffen Durinck and Wolfgang Huber provide an powerful interface between
the R language and [Ensembl Biomart](http://ensemblgenomes.org/info/access/biomart) by providing the R package [biomaRt](http://www.bioconductor.org/packages/release/bioc/html/biomaRt.html).
However, the `biomartr` package extends the functionality of the [biomaRt](http://www.bioconductor.org/packages/release/bioc/html/biomaRt.html) package
and introduces a more organism centered annotation retrieval concept.

The following sections will introduce users to the functionality and data retrieval precedures of `biomartr` and will show how `biomartr`
extends the functionality of the initial [biomaRt](http://www.bioconductor.org/packages/release/bioc/html/biomaRt.html) package.

### Getting Started with biomaRt

The best way to get started with the methodology presented by the established [biomaRt](http://www.bioconductor.org/packages/release/bioc/html/biomaRt.html) package is to understand the workflow
of data retrieval. The database provided by [Ensembl Biomart](http://ensemblgenomes.org/info/access/biomart) is organized in so called:
`marts`, `datasets`, and `attributes`. So when users want to retrieve information for a specific organism
of interest, first they need to specify the `marts` and `datasets` in which the information of the corresponding
organism can be found. Subsequently they can specify the `attributes` argument that is ought to be returned for the corresponding organism.

The availability of `marts`, `datasets`, and `attributes` can be checked by the following functions:

```{r,eval=TRUE}
# install the biomaRt package
# source("http://bioconductor.org/biocLite.R")
# biocLite("biomaRt")

# load biomaRt
library(biomaRt)

# look at top 10 databases
head(listMarts(host = "www.ensembl.org"), 10)
```

Users will observe that several `marts` providing annotation for
specific classes of organisms or groups of organisms are available.

For our example, we will choose the `hsapiens_gene_ensembl` `mart` and list all available
datasets that are element of this `mart`.

```{r,eval=TRUE}
head(listDatasets(useMart("ENSEMBL_MART_ENSEMBL", host = "www.ensembl.org")), 10)
```

The `useMart()` function is a wrapper function provided by `biomaRt` to connect a selected BioMart
database (`mart`) with a corresponding dataset stored within this `mart`.

We select dataset `hsapiens_gene_ensembl` and now check for available attributes (annotation data)
that can be accessed for `Homo sapiens` genes.

```{r,eval=TRUE}
head(listAttributes(useDataset(dataset = "hsapiens_gene_ensembl", 
                               mart    = useMart("ENSEMBL_MART_ENSEMBL",
                               host    = "www.ensembl.org"))), 10)
```

Please note the nested structure of this attribute query. For an attribute query procedure
an additional wrapper function named `useDataset()` is needed in which `useMart()` and a corresponding 
dataset needs to be specified. The result is a table storing the name of available attributes for
_Homo sapiens_ as well as a short description.

Furthermore, users can retrieve all filters for _Homo sapiens_ that can be specified by the actual BioMart query process.

```{r,eval=TRUE}
head(listFilters(useDataset(dataset = "hsapiens_gene_ensembl", 
                            mart    = useMart("ENSEMBL_MART_ENSEMBL",
                            host    = "www.ensembl.org"))), 10)
```

After accumulating all this information, it is now possible to perform an actual BioMart query
by using the `getBM()` function.

In this example we will retrieve attributes: `start_position`,`end_position` and `description`
for the _Homo sapiens_ gene `"GUCA2A"`.

Since the input genes are `ensembl gene ids`, we need to specify the `filters` argument `filters = "hgnc_symbol"`.

```{r,eval=TRUE}
# 1) select a mart and data set
mart <- useDataset(dataset = "hsapiens_gene_ensembl", 
                   mart    = useMart("ENSEMBL_MART_ENSEMBL",
                   host    = "www.ensembl.org"))

# 2) run a biomart query using the getBM() function
# and specify the attributes and filter arguments
geneSet <- "GUCA2A"

resultTable <- getBM(attributes = c("start_position","end_position","description"),
                     filters    = "hgnc_symbol", 
                     values     = geneSet, 
                     mart       = mart)

resultTable 
```

When using `getBM()` users can pass all attributes retrieved by `listAttributes()` to the `attributes` argument
of the `getBM()` function.

# Getting Started with `biomartr`

This query methodology provided by `Ensembl Biomart` and the `biomaRt` package is a very well defined approach
for accurate annotation retrieval. Nevertheless, when learning this query methodology it (subjectively)
seems non-intuitive from the user perspective. Therefore, the `biomartr` package provides another
query methodology that aims to be more organism centric.

Taken together, the following workflow allows users to perform fast BioMart queries for 
attributes using the `biomart()` function implemented in this `biomartr` package:

1) get attributes, datasets, and marts via : `organismAttributes()`

2) choose available biological features (filters) via: `organismFilters()`

3) specify a set of query genes: e.g. retrieved with `getGenome()`, `getProteome()` or `getCDS()`

4) specify all arguments of the `biomart()` function using steps 1) - 3) and
perform a BioMart query

__Note that dataset names change very frequently due to the update of dataset versions.
So in case some query functions do not work properly, users should check with
`organismAttributes(update = TRUE)` whether or not their dataset name has been changed.
For example, `organismAttributes("Homo sapiens", topic = "id", update = TRUE)`
might reveal that the dataset `ENSEMBL_MART_ENSEMBL` has changed.__


## Retrieve marts, datasets, attributes, and filters with biomartr

### Retrieve Available Marts

The `getMarts()` function allows users to list all available databases that can be accessed through BioMart interfaces.

```{r,eval=TRUE}
# load the biomartr package
library(biomartr)

# list all available databases
getMarts()
```

### Retrieve Available Datasets from a Specific Mart

Now users can select a specific database to list all available datasets
that can be accessed through this database. In this example we choose
the `ENSEMBL_MART_ENSEMBL` database.

```{r,eval=TRUE}
head(getDatasets(mart = "ENSEMBL_MART_ENSEMBL") , 5)
```

Now you can select the dataset `hsapiens_gene_ensembl` and list all available attributes that can be retrieved from this dataset.

```{r,eval=TRUE}
tail(getDatasets(mart = "ENSEMBL_MART_ENSEMBL") , 38)
```

### Retrieve Available Attributes from a Specific Dataset

Now that you have selected a database (`hsapiens_gene_ensembl`) and a dataset (`hsapiens_gene_ensembl`),
users can list all available attributes for this dataset using the `getAttributes()` function.

```{r,eval=TRUE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# list all available attributes for dataset: hsapiens_gene_ensembl
head( getAttributes(mart    = "ENSEMBL_MART_ENSEMBL", 
                    dataset = "hsapiens_gene_ensembl"), 10 )
```


### Retrieve Available Filters from a Specific Dataset

Finally, the `getFilters()` function allows users to list available filters
for a specific dataset that can be used for a `biomart()` query.

```{r,eval=TRUE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# list all available filters for dataset: hsapiens_gene_ensembl
head( getFilters(mart    = "ENSEMBL_MART_ENSEMBL", 
                 dataset = "hsapiens_gene_ensembl"), 10 )
```


## Organism Specific Retrieval of Information

In most use cases, users will work with a single or a set of model organisms. In this process they will mostly be
interested in specific annotations for this particular model organism. The `organismBM()`
function addresses this issue and provides users with an organism centric query to `marts` and `datasets`
which are available for a particular organism of interest.


__Note__ that when running the following functions for the first time, the data retrieval procedure will take some time, due to the remote access to BioMart. The corresponding result is then saved in a `*.txt` file named `_biomart/listDatasets.txt` within the `tempdir()` folder, allowing subsequent queries to be performed much faster.
The `tempdir()` folder, however, will be deleted after a new R session was established. In this case
the inital call of the subsequent functions again will take time to retrieve all organism specific data from the BioMart database.

This concept of locally storing all organism specific database linking information available in BioMart into
an internal file allows users to significantly speed up subsequent retrieval queries for that particular organism.


```{r,eval=TRUE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# retrieving all available datasets and biomart connections for
# a specific query organism (scientific name)
organismBM(organism = "Homo sapiens")
```


The result is a table storing all `marts` and `datasets` from which annotations can be retrieved
for _Homo sapiens_. Furthermore, a short description as well as the version of the dataset
being accessed (very useful for publications) is returned.

Users will observe that 3 different `marts` provide 6 different `datasets` storing annotation information for
_Homo sapiens_.

__Please note, however, that scientific names of organisms must be written correctly! For ex. "Homo Sapiens" will be treated differently (not recognized) than "Homo sapiens" (recognized).__

Similar to the `biomaRt` package query methodology, users need to specify `attributes` and `filters` to be able to perform
accurate BioMart queries. Here the functions `organismAttributes()` and `organismFilters()` provide useful and intuitive
concepts to obtain this information.


```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# return available attributes for "Homo sapiens"
head(biomartr::organismAttributes("Homo sapiens"), 20)
```

Users will observe that the `organismAttributes()` function returns a data.frame storing attribute names, datasets, and marts which
are available for `Homo sapiens`. After the ENSEMBL release 87 the `ENSEMBL_MART_SEQUENCE` service provided
by Ensembl does not work properly and thus the `organismAttributes()` function prints out warning messages to make the user aware
when certain marts provided bt Ensembl do not work properly, yet. 

An additional feature provided by `organismAttributes()` is the `topic` argument. The `topic` argument allows users to to search for specific attributes,  topics, or categories for faster filtering.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for attribute topic "id"
head(organismAttributes("Homo sapiens", topic = "id"), 20)
```

Now, all `attribute names` having `id` as part of their `name` are being returned.

Another example is `topic = "homolog"`.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for attribute topic "homolog"
head(organismAttributes("Homo sapiens", topic = "homolog"), 20)
```

Or `topic = "dn"` and `topic = "ds"` for `dn` and `ds` value retrieval.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for attribute topic "dn"
head(organismAttributes("Homo sapiens", topic = "dn"))
```


```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for attribute topic "ds"
head(organismAttributes("Homo sapiens", topic = "ds"))
```

Analogous to the `organismAttributes()` function, the `organismFilters()` function returns
all filters that are available for a query organism of interest.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# return available filters for "Homo sapiens"
head(organismFilters("Homo sapiens"), 20)
```

The `organismFilters()` function also allows users to search for filters that correspond to
a specific topic or category.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for filter topic "id"
head(organismFilters("Homo sapiens", topic = "id"), 20)
```


## Performing BioMart queries with biomartr

The short introduction to the functionality of
`organismBM()`, `organismAttributes()`, and `organismFilters()`
will allow users to perform BioMart queries in a very intuitive 
organism centric way. The main function to perform BioMart queries
is `biomart()`.


For the following examples we will assume that we are interested in the annotation of specific genes from the _Homo sapiens_ proteome. We want to map the corresponding refseq gene id to a set of other gene ids used in other databases. For this purpose, first we need consult the `organismAttributes()` function.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)

head(organismAttributes("Homo sapiens", topic = "id"))
```

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# retrieve the proteome of Homo sapiens from refseq
file_path <- getProteome( db       = "refseq",
                          organism = "Homo sapiens",
                          path     = file.path("_ncbi_downloads","proteomes") )

Hsapiens_proteome <- read_proteome(file_path, format = "fasta")

# remove splice variants from id
gene_set <- unlist(sapply(strsplit(Hsapiens_proteome@ranges@NAMES[1:5], ".",fixed = TRUE), function(x) x[1]))

result_BM <- biomart( genes      = gene_set,
                      mart       = "ENSEMBL_MART_ENSEMBL", 
                      dataset    = "hsapiens_gene_ensembl",
                      attributes = c("ensembl_gene_id","ensembl_peptide_id"),
                      filters    = "refseq_peptide")

result_BM 
```


The `biomart()` function takes as arguments a set of genes (gene ids specified in the `filter` argument), the corresponding `mart` and `dataset`, as well as the `attributes` which shall be returned.

## Gene Ontology 

The `biomartr` package also enables a fast and intuitive retrieval of GO terms
and additional information via the `getGO()` function. Several databases can be selected
to retrieve GO annotation information for a set of query genes. So far, the `getGO()`
function allows GO information retrieval from the [Ensembl Biomart](http://ensemblgenomes.org/info/access/biomart) database. 

In this example we will retrieve GO information for a set of _Homo sapiens_ genes
stored as `hgnc_symbol`.

### GO Annotation Retrieval via BioMart

The `getGO()` function takes several arguments as input to retrieve GO information from BioMart. 
First, the scientific name of the `organism` of interest needs to be specified. Furthermore, a set of
`gene ids` as well as their corresponding `filter` notation (`GUCA2A` gene ids have `filter` notation `hgnc_symbol`; see `organismFilters()` for details)
need to be specified. The `database` argument then defines the database from which GO information shall be retrieved.

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
# search for GO terms of an example Homo sapiens gene
GO_tbl <- getGO(organism = "Homo sapiens", 
                genes    = "GUCA2A",
                filters  = "hgnc_symbol")

GO_tbl
```


Hence, for each _gene id_ the resulting table stores all annotated GO terms found in [Ensembl Biomart](http://ensemblgenomes.org/info/access/biomart).

