---
title: "Perform Phylotranscriptomics with `biomartr`"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phylotranscriptomics using biomartr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(width = 750)
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
```

## Introduction

`Phylotranscriptomics` is a sub-field of Evolutionary Developmental Biology research investigating the evolutionary conservation of transcriptomes during development. The term `Phylotranscriptomics` defines the combination of evolutionary information with transcriptome data to capture signals of evolutionary conservation.

The [Introduction to Phylotranscriptomics](https://github.com/HajkD/myTAI/blob/master/vignettes/Introduction.Rmd) vignette provided by the [myTAI](https://github.com/HajkD/myTAI) package
will give you a broad overwiew of phylotranscriptomics as scientific subject.

The `biomartr` package aims to provide an interface for functional annotation
queries. Hence, combining functional annotation with the detection of evolutionary signals in
biological processes with [orthologr](https://github.com/HajkD/orthologr) and [myTAI](https://github.com/HajkD/myTAI) allows users to correlate 
biological functions and constraints with their evolutionary origin.

This vignette aims to provide use cases for combining functional annotation
of genes (`biomartr`) with their role in development (`myTAI`) in an evolutionary context (`orthologr`).


## Use Case #1: Functional Annotation of Genes Sharing a Common Evolutionary History

Phylotranscriptomics aims to predict stages or periods of evolutionary conservation in
biological processes on the transcriptome level. However, finding genes sharing a [common evolutionary history](https://github.com/HajkD/myTAI/blob/master/vignettes/Enrichment.Rmd) could reveal how the the biological process might have evolved in the first place.

In this `Use Case` we will combine functional and biological annotation obtained with `biomartr` with enriched genes obtained with [PlotEnrichment()](https://github.com/HajkD/myTAI/blob/master/vignettes/Enrichment.Rmd). 

### Step 1

Install and load the [myTAI](https://github.com/HajkD/myTAI) package:

```{r, eval=FALSE}
# install myTAI
install.packages("myTAI")

# load myTAI
library(myTAI)

```




