---
title: "Sequence Retrieval"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sequence Retrieval}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
options(width = 750)
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
```


## Biological Sequence Retrieval 

The `biomartr` package allows users to retrieve biological sequences in a very simple and intuitive way.

Using `biomartr`, users can retrieve either genomes, proteomes, CDS, RNA, GFF, and genome assembly statistics data using the specialized functions:

* [1. Genome retrieval with `getGenome()`](#genome-retrieval)
    - [1.1 from NCBI RefSeq](#example-ncbi-refseq)
    - [1.2 from NCBI Genbank](#example-ncbi-genbank)
    - [1.3 from ENSEMBL](#example-ensembl)
    - [1.4 from ENSEMBLGENOMES](#example-ensemblgenomes)
* [2. Proteome retrieval with `getProteome()`](#proteome-retrieval)
    - [2.1 from NCBI RefSeq](#example-ncbi-refseq-1)
    - [2.2 from NCBI Genbank](#example-ncbi-genbank-1)
    - [2.3 from ENSEMBL](#example-ensembl-1)
    - [2.4 from ENSEMBLGENOMES](#example-ensemblgenomes-1)
    - [2.5 from UniProt](#example-uniprot)
* [3. Coding sequence retrieval with `getCDS()`](#cds-retrieval)
    - [3.1 from NCBI RefSeq](#example-ncbi-refseq-2)
    - [3.2 from NCBI Genbank](#example-ncbi-genbank-2)
    - [3.3 from ENSEMBL](#example-ensembl-2)
    - [3.4 from ENSEMBLGENOMES](#example-ensemblgenomes-2)
* [4. RNA retrieval with `getRNA()`](#rna-retrieval)
    - [4.1 from NCBI RefSeq](#example-ncbi-refseq-3)
    - [4.2 from NCBI Genbank](#example-ncbi-genbank-3)
    - [4.3 from ENSEMBL](#example-ensembl-3)
    - [4.4 from ENSEMBLGENOMES](#example-ensemblgenomes-3)
* [5. GFF retrieval with `getGFF()`](#retrieve-the-annotation-file-of-a-particular-genome)
    - [5.1 from NCBI RefSeq](#example-ncbi-refseq-4)
    - [5.2 from NCBI Genbank](#example-ncbi-genbank-4)
    - [5.3 from ENSEMBL](#example-ensembl-4)
    - [5.4 from ENSEMBLGENOMES](#example-ensemblgenomes-4)
* [6. GTF retrieval with `getGTF()`](#retrieve-the-annotation-file-of-a-particular-genome)
    - [6.1 from NCBI RefSeq](#example-ncbi-refseq-5)
    - [6.2 from NCBI Genbank](#example-ncbi-genbank-5)
    - [6.3 from ENSEMBL](#example-ensembl-5)
    - [6.4 from ENSEMBLGENOMES](#example-ensemblgenomes-5)
* [7. Repeat Masker Annotation file retrieval with `getRepeatMasker()`](#repeat-masker-retrieval)
    - [7.1 from NCBI RefSeq](#example-ncbi-refseq-6)
    

## Getting Started with Sequence Retrieval

First users can check whether or not the genome, proteome, CDS, RNA, GFF, GTF, or genome assembly statistics of their interest is available for download.

Using the scientific name of the organism of interest, users can check whether the 
corresponding genome is available via the `is.genome.available()` function.

Please note that the first time you run this command it might take a while, because
during the initial execution of this function all necessary information is retrieved from NCBI
and then stored locally. All further runs are then much faster.

Example `refseq`: 

```r
# checking whether or not the Homo sapiens 
# genome is avaialable for download
is.genome.available("Homo sapiens", db = "refseq")
```

```
[1] TRUE
```

Example `genbank`: 

```r
# checking whether or not the Homo sapiens 
# genome is avaialable for download
is.genome.available("Homo sapiens", db = "genbank")
```

```
[1] TRUE
```

### Using `is.genome.available()` with ENSEMBL and ENSEMBLGENOMES

Users can also specify `db = "ensembl"` or `db = "ensemblgenomes"` to retrieve available organisms provided by ENSEMBL or ENSEMBLGENOMES.
Again, users might experience a delay in the execution of this function when running it for the first time.
This is due to the download of ENSEMBL or ENSEMBLGENOMES information which is then stored internally to enable a much faster
execution of this function in following runs. The corresponding information files are stored at `file.path(tempdir(), "ensembl_summary.txt")` 
and `file.path(tempdir(), "ensemblgenomes_summary.txt")`.


Example `ENSEMBL`: 

```{r,eval=FALSE}
# cheking whether Homo sapiens is available in the ENSEMBL database
is.genome.available("Homo sapiens", db = "ensembl")
```

```
[1] TRUE
```

```{r,eval=FALSE}
# retrieve details for Homo sapiens
is.genome.available("Homo sapiens", db = "ensembl", details = TRUE)
```

```
  division taxon_id         name release display_name        accession common_name
1  Ensembl     9606 homo_sapiens      86        Human GCA_000001405.22       human
                                                                    aliases
1 homo, homo sapiens, h_sapiens, enshs, human, hsap, 9606, homsap, hsapiens
                                                       groups assembly
1 core, cdna, vega, otherfeatures, rnaseq, variation, funcgen   GRCh38
```

Example `ENSEMBLGENOMES`: 

For example, some species that cannot be found at `db = "ensembl"` might be available at `db = "ensemblgenomes"`
and vice versa. So I recommend users to to check in both databases ENSEMBL and ESEMBLGENOMES whether or not a
particular species is present. In case of `"Homo sapiens"`, the genome is available at `db = "ensembl"` but
__not__ at `db = "ensemblgenomes"` whereas the genome of `"Arabidopsis thaliana"` is available at `db = "ensemblgenomes"`
but __not__ at `db = "ensembl"`.


```{r,eval=FALSE}
# cheking whether Homo sapiens is available in the ENSEMBLGENOMES database
is.genome.available("Homo sapiens", db = "ensemblgenomes")
    ```

```
Error: Unfortunately organism 'Homo sapiens' is not available at ENSEMBLGENOMES. 
Please check whether or not the organism name is typed correctly.
```

```{r,eval=FALSE}
# cheking whether Arabidopsis thaliana is available in the ENSEMBLGENOMES database
is.genome.available("Arabidopsis thaliana", db = "ensemblgenomes")
```

```
[1] TRUE
```

```{r,eval=FALSE}
# show details for Arabidopsis thaliana
is.genome.available("Arabidopsis thaliana", db = "ensemblgenomes", details = TRUE)
```

```
       division taxon_id                 name release         display_name
          <chr>    <int>                <chr>   <int>                <chr>
1 EnsemblPlants     3702 arabidopsis_thaliana      85 Arabidopsis thaliana
```

Please note that the detailed information provided by ENSEMBL or ENSEMBL genomes differs from the information provided by NCBI. 


By specifying the `details = TRUE` argument, the genome file size as well as 
additional information can be printed to the console.

```{r,eval=FALSE}
# printing details to the console
is.genome.available("Homo sapiens", details = TRUE, db = "refseq")
```

```
  assembly_accession  bioproject    biosample     wgs_master
               <chr>       <chr>        <chr>          <chr>
1   GCF_000001405.35    PRJNA168         <NA>           <NA>
2    GCF_000306695.2 PRJNA178030 SAMN02205338 AMYH00000000.2
 with 25 more variables: refseq_category <chr>, taxid <int>,
   species_taxid <int>, organism_name <chr>, infraspecific_name <chr>,
   isolate <chr>, version_status <chr>, assembly_level <chr>,
   release_type <chr>, genome_rep <chr>, seq_rel_date <date>,
   asm_name <chr>, submitter <chr>, gbrs_paired_asm <chr>,
   paired_asm_comp <chr>, ftp_path <chr>, excluded_from_refseq <chr>,
   kingdoms <chr>, group <chr>, subgroup <chr>, file_size_MB <dbl>,
   chrs <int>, organelles <int>, plasmids <int>, bio_projects <int>
```

The argument `db` specifies from which database (`refseq`, `genbank`, `ensembl` or `ensemblgenomes`) organism
information shall be retrieved.

Users can determine the total number of available genomes using the `listGenomes()` function.

Example `refseq`: 

```{r,eval=FALSE}
length(listGenomes(db = "refseq"))
```

```
[1] 13689
```

Example `genbank`: 

```{r,eval=FALSE}
length(listGenomes(db = "genbank"))
```

```
[1] 20484
```

Example `ensembl`: 

```{r,eval=FALSE}
length(listGenomes(db = "ensembl"))
```

```
[1] 112
```

Example `ensemblgenomes`: 

```{r,eval=FALSE}
length(listGenomes(db = "ensemblgenomes"))
```

```
[1] 45170
```

Hence, currently 13689 genomes (including all kingdoms of life) are stored on NCBI RefSeq (as of 07/02/2018).

### Retrieving kingdom, group and subgroup information

Using this example users can retrieve the number of available species for each kingdom of life:

Example `refseq`:

```{r,eval=FALSE}
# the number of genomes available for each kingdom
listKingdoms(db = "refseq")
```

```
    Archaea  Bacteria Eukaryota   Viroids   Viruses 
      163      5777       541        47      7161
```

Example `genbank`:

```{r,eval=FALSE}
# the number of genomes available for each kingdom
listKingdoms(db = "genbank")
```

```
  Archaea  Bacteria Eukaryota 
     1129     17235      2120
```

Example `ENSEMBL`:

```{r,eval=FALSE}
# the number of genomes available for each kingdom
listKingdoms(db = "ensembl")
```

```
Ensembl 
    112
```

Example `ENSEMBLGENOMES`:

```{r,eval=FALSE}
# the number of genomes available for each kingdom
listKingdoms(db = "ensemblgenomes")
```

```
EnsemblBacteria    EnsemblFungi  EnsemblMetazoa   EnsemblPlants EnsemblProtists 
          44048             811              69              53             189
```

### Analogous computations can be performed for `groups` and `subgroups`

Unfortunately, `ENSEMBL` and `ENSEMBLGENOMES` do not provide group or subgroup information.
Therefore, group and subgroup listings are limited to `refseq` and `genbank`.


Example `refseq`:

```{r,eval=FALSE}
# the number of genomes available for each group
listGroups(db = "refseq")
```

```
                Acidobacteria                       Animals 
                           11                           293 
                Avsunviroidae                    Deltavirus 
                            4                             1 
  dsDNA viruses, no RNA stage                 dsRNA viruses 
                         2572                           261 
                Elusimicrobia                 Euryarchaeota 
                            1                            64 
                    FCB group                         Fungi 
                          155                            34 
                 Fusobacteria                   Nitrospirae 
                            6                             3 
                        Other                        Plants 
                           10                            62 
                Pospiviroidae                Proteobacteria 
                           34                           774 
                     Protists                     PVC group 
                           34                            20 
   Retro-transcribing viruses                    Satellites 
                          135                           213 
                 Spirochaetes                 ssDNA viruses 
                           12                           864 
                ssRNA viruses                 Synergistetes 
                         1432                             1 
                   TACK group           Terrabacteria group 
                           13                           630 
        Thermodesulfobacteria                   Thermotogae 
                            3                             2 
           unassigned viruses          unclassified Archaea 
                            9                             1 
unclassified archaeal viruses         unclassified Bacteria 
                            3                             9 
          unclassified phages          unclassified viroids 
                           23                             8 
      unclassified virophages          unclassified viruses 
                            6                           176  
```

```{r,eval=FALSE}
# the number of genomes available for each subgroup
head(listSubgroups(db = "refseq"), 15)
```

```
         Acidithiobacillia             Acidobacteriia 
                         2                          8 
            Actinobacteria               Adenoviridae 
                       194                         55 
         Alloherpesviridae          Alphaflexiviridae 
                         7                         48 
       Alphaproteobacteria          Alphatetraviridae 
                       256                          3 
            Alvernaviridae              Amalgaviridae 
                         1                          4 
                Amphibians             Ampullaviridae 
                         3                          3 
             Anelloviridae              Apicomplexans 
                        53                         16 
Apple fruit crinkle viroid 
                         1 
```

Example `genbank`:

```{r,eval=FALSE}
# the number of genomes available for each group
listGroups(db = "genbank")
```

```
                   Acidobacteria                         Animals 
                             47                             655 
                    Caldiserica                 Deferribacteres 
                              1                               1 
                    DPANN group                   Elusimicrobia 
                             27                              57 
          environmental samples                   Euryarchaeota 
                              6                             197 
                      FCB group                           Fungi 
                            545                             508 
                   Fusobacteria Nitrospinae/Tectomicrobia group 
                              7                              11 
                    Nitrospirae                           Other 
                             46                               7 
                         Plants                  Proteobacteria 
                            193                            1490 
                       Protists                       PVC group 
                            142                             180 
                   Spirochaetes                   Synergistetes 
                             26                              12 
                     TACK group             Terrabacteria group 
                             73                            1072 
          Thermodesulfobacteria                     Thermotogae 
                              3                               9 
           unclassified Archaea           unclassified Bacteria 
                             48                            1407
```

```{r,eval=FALSE}
# the number of genomes available for each subgroup
head(listSubgroups(db = "genbank"), 15)
```

```
           Acidithiobacillia               Acidobacteriia 
                           5                            9 
              Actinobacteria          Alphaproteobacteria 
                         313                          411 
                  Amphibians                Apicomplexans 
                           4                           32 
     Archaea candidate phyla                 Archaeoglobi 
                          25                            3 
             Armatimonadetes                  Ascomycetes 
                          10                          361 
    Bacteria candidate phyla Bacteroidetes/Chlorobi group 
                        1372                          405 
              Basidiomycetes           Betaproteobacteria 
                         122                          268 
                       Birds 
                          59 
```


__Note__ that when running the `listGenomes()` function for the first time, it might take a while until
the function returns any results, because necessary information need to be downloaded from NCBI and ENSEMBL databases. All subsequent executions of `listGenomes()`
will then respond very fast, because they will access the corresponding files stored on your hard drive.

## Downloading Biological Sequences and Annotations

After checking for the availability of sequence information for an organism of interest,
the next step is to download the corresponding genome, proteome, CDS, or GFF file.
The following functions allow users to download proteomes, genomes, CDS and GFF files from several
database resources such as: [NCBI RefSeq](http://www.ncbi.nlm.nih.gov/refseq/about/), [NCBI Genbank](http://www.ncbi.nlm.nih.gov/genbank/about/), [ENSEMBL](http://www.ensembl.org/index.html) and [ENSEMBLGENOMES](http://ensemblgenomes.org/). When a corresponding proteome, genome, CDS or GFF file was
loaded to your hard-drive, a documentation `*.txt` file is generated storing `File Name`, `Organism`,
`Database`, `URL`, `DATE`, `assembly_accession`, `bioproject`, `biosample`, `taxid`, `version_status`, `release_type`,
`seq_rel_date` etc. information of the retrieved file. This way a better reproducibility of proteome, genome, CDS and GFF versions
used for subsequent data analyses can be achieved.


### Genome Retrieval

The easiest way to download a genome is to use the `getGenome()` function.

In this example we will download the genome of `Homo sapiens`.

The `getGenome()` function is an interface function to the [NCBI RefSeq](http://www.ncbi.nlm.nih.gov/refseq/about/), [NCBI Genbank](http://www.ncbi.nlm.nih.gov/genbank/about/),
[ENSEMBL](http://www.ensembl.org/index.html) and [ENSEMBLGENOMES](http://ensemblgenomes.org/) databases from
which corresponding genomes can be retrieved.

The `db` argument specifies from which database genome assemblies in `*.fasta` file format shall be retrieved.

Options are:

- `db = "refseq"` for retrieval from [NCBI RefSeq](http://www.ncbi.nlm.nih.gov/refseq/about/)
- `db = "genbank"` for retrieval from [NCBI Genbank](http://www.ncbi.nlm.nih.gov/genbank/about/)
- `db = "ensembl"` for retrieval from [ENSEMBL](http://www.ensembl.org/index.html)
- `db = "ensemblgenomes"` for retrieval from [ENSEMBLGENOMES](http://ensemblgenomes.org/)

Furthermore, users need to specify the scientific name of the organism of interest for which a genome assembly shall be downloaded, e.g. `organism = "Homo sapiens"`.
Finally, the `path` argument specifies the folder path in which the corresponding assembly shall be locally stored. In case users would like to store the genome file at a different location,
they can specify the `path = file.path("put","your","path","here")` argument.

### Example NCBI RefSeq:

```{r,eval=FALSE}
# download the genome of Homo sapiens from refseq
# and store the corresponding genome file in '_ncbi_downloads/genomes'
HS.genome.refseq <- getGenome( db       = "refseq", 
                               organism = "Homo sapiens",
                               path     = file.path("_ncbi_downloads","genomes") )
```

In this example, `getGenome()` creates a directory named `'_ncbi_downloads/genomes'` into which the corresponding genome named `GCF_000001405.34_GRCh38.p8_genomic.fna.gz` is downloaded. The return value of `getGenome()` is the folder path to the downloaded genome file
that can then be used as input to the `read_genome()` function. The variable `HS.genome.refseq` stores the path to the downloaded genome. 

Users can also omit the `path` argument if they wish to store the genome in their current working directory. E.g.:

```{r,eval=FALSE}
# download the genome of Homo sapiens from refseq
# and store the corresponding genome file in '_ncbi_downloads/genomes'
HS.genome.refseq <- getGenome( db       = "refseq", 
                               organism = "Homo sapiens")
```


Subsequently, users can use the `read_genome()` function to import the genome into the R session. Users can choose to work with the genome sequence in R either as [Biostrings](http://bioconductor.org/packages/release/bioc/html/Biostrings.html) object (`obj.type = "Biostrings"`) or [data.table](https://github.com/Rdatatable/data.table/wiki) object
(`obj.type = "data.table"`) by specifying the `obj.type` argument of the `read_genome()` function. 

```{r,eval=FALSE}
# import downloaded genome as Biostrings object
Human_Genome <- read_genome(file     = HS.genome.refseq, 
                            obj.type = "Biostrings")

```

```{r,eval=FALSE}
# look at the Biostrings object
Human_Genome
```

```
  A DNAStringSet instance of length 551
          width seq                                                names               
  [1] 248956422 NNNNNNNNNNNNNNNNNNNNNNNN...NNNNNNNNNNNNNNNNNNNNNNN NC_000001.11 Homo...
  [2]    175055 GAATTCAGCTGAGAAGAACAGGCA...TGTTTGTCAGTCACATAGAATTC NT_187361.1 Homo ...
  [3]     32032 AGGGGTCTGCTTAGAGAGGGTCTC...TGACTTACGTTGACGTGGAATTC NT_187362.1 Homo ...
  [4]    127682 GATCGAGACTATCCTGGCTAACAC...ATTGTCAATTGGGACCTTTGATC NT_187363.1 Homo ...
  [5]     66860 GAATTCATTCGATGACGATTCCAT...AAAAAACTCTCAGCCACGAATTC NT_187364.1 Homo ...
  ...       ... ...
[547]    170148 TTTCTTTCTTTTTTTTTTTTTTGT...GTCACAGGACTCATGGGGAATTC NT_187685.1 Homo ...
[548]    215732 TGTGGTGAGGACCCTTAAGATCTA...GTCACAGGACTCATGGGGAATTC NT_187686.1 Homo ...
[549]    170537 TCTACTCTCCCATGCTTGCCTCGG...GTCACAGGACTCATGGGGAATTC NT_187687.1 Homo ...
[550]    177381 GATCTATCTGTATCTCCACAGGTG...GTCACAGGACTCATGGGGAATTC NT_113949.2 Homo ...
[551]     16569 GATCACAGGTCTATCACCCTATTA...CCCTTAAATAAGACATCACGATG NC_012920.1 Homo ...
```

Internally, a text file named `doc_Homo_sapiens_db_refseq.txt` is generated. The information stored in this log file is structured as follows:

```
File Name: Homo_sapiens_genomic_refseq.fna.gz
Organism Name: Homo_sapiens
Database: NCBI refseq
URL: ftp://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/001/405/
GCF_000001405.35_GRCh38.p9/GCF_000001405.35_GRCh38.p9_genomic.fna.gz
Download_Date: Sat Oct 22 12:41:07 2016
refseq_category: reference genome
assembly_accession: GCF_000001405.35
bioproject: PRJNA168
biosample: NA
taxid: 9606
infraspecific_name: NA
version_status: latest
release_type: Patch
genome_rep: Full
seq_rel_date: 2016-09-26
submitter: Genome Reference Consortium
```

In summary, the `getGenome()` and `read_genome()` functions allow users to retrieve genome assemblies by specifying
the scientific name of the organism of interest and allow them to import the retrieved genome assembly e.g. as `Biostrings` object.
Thus, users can then perform the `Biostrings notation` to work with downloaded genomes and can rely on the log
file generated by `getGenome()` to better document the source and version of genome assemblies used for subsequent studies.

Alternatively, users can perform the pipeline logic of the [magrittr](https://github.com/smbache/magrittr) package:

```{r,eval=FALSE}
# install.packages("magrittr")
library(magrittr)
# import genome as Biostrings object
Human_Genome <- getGenome( db       = "refseq", 
                           organism = "Homo sapiens",
                           path     = file.path("_ncbi_downloads","genomes")) %>%
    read_genome(obj.type = "Biostrings")
```

```{r,eval=FALSE}
Human_Genome
```

```
  A DNAStringSet instance of length 551
          width seq                                                names               
  [1] 248956422 NNNNNNNNNNNNNNNNNNNNNNNN...NNNNNNNNNNNNNNNNNNNNNNN NC_000001.11 Homo...
  [2]    175055 GAATTCAGCTGAGAAGAACAGGCA...TGTTTGTCAGTCACATAGAATTC NT_187361.1 Homo ...
  [3]     32032 AGGGGTCTGCTTAGAGAGGGTCTC...TGACTTACGTTGACGTGGAATTC NT_187362.1 Homo ...
  [4]    127682 GATCGAGACTATCCTGGCTAACAC...ATTGTCAATTGGGACCTTTGATC NT_187363.1 Homo ...
  [5]     66860 GAATTCATTCGATGACGATTCCAT...AAAAAACTCTCAGCCACGAATTC NT_187364.1 Homo ...
  ...       ... ...
[547]    170148 TTTCTTTCTTTTTTTTTTTTTTGT...GTCACAGGACTCATGGGGAATTC NT_187685.1 Homo ...
[548]    215732 TGTGGTGAGGACCCTTAAGATCTA...GTCACAGGACTCATGGGGAATTC NT_187686.1 Homo ...
[549]    170537 TCTACTCTCCCATGCTTGCCTCGG...GTCACAGGACTCATGGGGAATTC NT_187687.1 Homo ...
[550]    177381 GATCTATCTGTATCTCCACAGGTG...GTCACAGGACTCATGGGGAATTC NT_113949.2 Homo ...
[551]     16569 GATCACAGGTCTATCACCCTATTA...CCCTTAAATAAGACATCACGATG NC_012920.1 Homo ...
```

### Example NCBI Genbank:

```{r,eval=FALSE}
# download the genome of Homo sapiens from Genbank
# and store the corresponding genome file in '_ncbi_downloads/genomes'
HS.genome.genbank <- getGenome( db       = "genbank", 
                                organism = "Homo sapiens",
                                path     = file.path("_ncbi_downloads","genomes") )
```

```{r,eval=FALSE}
# import downloaded genome as Biostrings object
Human_Genome <- read_genome(file     = HS.genome.genbank, 
                            obj.type = "Biostrings")
```

```{r,eval=FALSE}
# look at the Biostrings object
Human_Genome
```

```
  A DNAStringSet instance of length 551
          width seq                                                names               
  [1] 248956422 NNNNNNNNNNNNNNNNNNNNNNNN...NNNNNNNNNNNNNNNNNNNNNNN CM000663.2 Homo s...
  [2]    175055 GAATTCAGCTGAGAAGAACAGGCA...TGTTTGTCAGTCACATAGAATTC KI270706.1 Homo s...
  [3]     32032 AGGGGTCTGCTTAGAGAGGGTCTC...TGACTTACGTTGACGTGGAATTC KI270707.1 Homo s...
  [4]    127682 GATCGAGACTATCCTGGCTAACAC...ATTGTCAATTGGGACCTTTGATC KI270708.1 Homo s...
  [5]     66860 GAATTCATTCGATGACGATTCCAT...AAAAAACTCTCAGCCACGAATTC KI270709.1 Homo s...
  ...       ... ...
[547]    170148 TTTCTTTCTTTTTTTTTTTTTTGT...GTCACAGGACTCATGGGGAATTC KI270931.1 Homo s...
[548]    215732 TGTGGTGAGGACCCTTAAGATCTA...GTCACAGGACTCATGGGGAATTC KI270932.1 Homo s...
[549]    170537 TCTACTCTCCCATGCTTGCCTCGG...GTCACAGGACTCATGGGGAATTC KI270933.1 Homo s...
[550]    177381 GATCTATCTGTATCTCCACAGGTG...GTCACAGGACTCATGGGGAATTC GL000209.2 Homo s...
[551]     16569 GATCACAGGTCTATCACCCTATTA...CCCTTAAATAAGACATCACGATG J01415.2 Homo sap...
```


### Example ENSEMBL:

```{r,eval=FALSE}
# download the genome of Homo sapiens from ENSEMBL
# and store the corresponding genome file in '_ncbi_downloads/genomes'
HS.genome.ensembl <- getGenome( db       = "ensembl", 
                                organism = "Homo sapiens",
                                path     = file.path("_ncbi_downloads","genomes") )
```

```{r,eval=FALSE}
# import downloaded genome as Biostrings object
Human_Genome <- read_genome(file     = HS.genome.ensembl, 
                            obj.type = "Biostrings")
```

```{r,eval=FALSE}
# look at the Biostrings object
Human_Genome
```

```
  A DNAStringSet instance of length 524
          width seq                                                names               
  [1] 248956422 NNNNNNNNNNNNNNNNNNNNNNNN...NNNNNNNNNNNNNNNNNNNNNNN 1 dna:chromosome ...
  [2] 242193529 NNNNNNNNNNNNNNNNNNNNNNNN...NNNNNNNNNNNNNNNNNNNNNNN 2 dna:chromosome ...
  [3] 198295559 NNNNNNNNNNNNNNNNNNNNNNNN...NNNNNNNNNNNNNNNNNNNNNNN 3 dna:chromosome ...
  [4] 190214555 NNNNNNNNNNNNNNNNNNNNNNNN...NNNNNNNNNNNNNNNNNNNNNNN 4 dna:chromosome ...
  [5] 181538259 NNNNNNNNNNNNNNNNNNNNNNNN...NNNNNNNNNNNNNNNNNNNNNNN 5 dna:chromosome ...
  ...       ... ...
[520]       993 GCCCCACGTCCGGGAGGGAGGTGG...GAGGGAGGTGGGGGGTCAGCCCT KI270539.1 dna:sc...
[521]       990 TTTCATAGAGCATGTTTGAAACAC...TCAGAAACTTGTTGTGATGTGTG KI270385.1 dna:sc...
[522]       981 AGATTTCGTTGGAACGGGATAAAC...TCAGCTTTCAAACACTCTTTTTG KI270423.1 dna:sc...
[523]       971 ATTTGCGATGTGTGTTCTCAACTA...TTGGATAGCTTTGAAGTTTTCGT KI270392.1 dna:sc...
[524]       970 AAGTGGATATTTGGATAGCTTTGA...TCCTCAATAACAGAGTTGAACCT KI270394.1 dna:sc...
```

### Example ENSEMBLGENOMES:

Due to the unavailability of `"Homo sapiens"` at `db = "ensemblgenomes"` here we choose `"Arabidopsis thaliana"`
as example.

```{r,eval=FALSE}
# download the genome of Arabidopsis thaliana from ENSEMBLGENOMES
# and store the corresponding genome file in '_ncbi_downloads/genomes'
AT.genome.ensemblgenomes <- getGenome( db       = "ensemblgenomes", 
                                       organism = "Arabidopsis thaliana",
                                       path     = file.path("_ncbi_downloads","genomes") )
```

```{r,eval=FALSE}
# import downloaded genome as Biostrings object
Cress_Genome <- read_genome(file     = AT.genome.ensemblgenomes, 
                            obj.type = "Biostrings")
```

```{r,eval=FALSE}
# look at the Biostrings object
Cress_Genome
```

```
  A DNAStringSet instance of length 7
       width seq                                                   names               
[1] 30427671 CCCTAAACCCTAAACCCTAAACCCT...AGGGTTTAGGGTTTAGGGTTTAGGG 1 dna:chromosome ...
[2] 19698289 NNNNNNNNNNNNNNNNNNNNNNNNN...AGGGTTTAGGGTTTAGGGTTTAGGG 2 dna:chromosome ...
[3] 23459830 NNNNNNNNNNNNNNNNNNNNNNNNN...ACCCTAAACCCTAAACCCTAAACCC 3 dna:chromosome ...
[4] 18585056 NNNNNNNNNNNNNNNNNNNNNNNNN...AAGGGTTTAGGGTTTAGGGTTTAGG 4 dna:chromosome ...
[5] 26975502 TATACCATGTACCCTCAACCTTAAA...GTTTAGGATTTAGGGTTTTTAGATC 5 dna:chromosome ...
[6]   366924 GGATCCGTTCGAAACAGGTTAGCCT...TCGCAGAATGGAAACAAACCGGATT Mt dna:chromosome...
[7]   154478 ATGGGCGAACGACGGGAATTGAACC...TCATAATAACTTGGTCCCGGGCATC Pt dna:chromosome...
```

### Proteome Retrieval

The `getProteome()` function is an interface function to the [NCBI RefSeq](http://www.ncbi.nlm.nih.gov/refseq/about/), [NCBI Genbank](http://www.ncbi.nlm.nih.gov/genbank/about/),
[ENSEMBL](http://www.ensembl.org/index.html) and [ENSEMBLGENOMES](http://ensemblgenomes.org/) databases from
which corresponding proteomes can be retrieved. It works analogous to `getGenome()`. 

The `db` argument specifies from which database proteomes in `*.fasta` file format shall be retrieved.

Options are:

- `db = "refseq"` for retrieval from [NCBI RefSeq](http://www.ncbi.nlm.nih.gov/refseq/about/)
- `db = "genbank"` for retrieval from [NCBI Genbank](http://www.ncbi.nlm.nih.gov/genbank/about/)
- `db = "ensembl"` for retrieval from [ENSEMBL](http://www.ensembl.org/index.html)
- `db = "ensemblgenomes"` for retrieval from [ENSEMBLGENOMES](http://ensemblgenomes.org/)

Furthermore, again users need to specify the scientific name of the organism of interest for which a proteomes shall be downloaded, e.g. `organism = "Homo sapiens"`.
Finally, the `path` argument specifies the folder path in which the corresponding proteome shall be locally stored. In case users would like to store the proteome file at a different location,
they can specify the `path = file.path("put","your","path","here")` argument.

### Example `NCBI RefSeq`:

```{r,eval=FALSE}
# download the proteome of Homo sapiens from refseq
# and store the corresponding proteome file in '_ncbi_downloads/proteomes'
HS.proteome.refseq <- getProteome( db       = "refseq", 
                                   organism = "Homo sapiens",
                                   path     = file.path("_ncbi_downloads","proteomes"))
```

In this example, `getProteome()` creates a directory named `'_ncbi_downloads/proteomes'` into which the corresponding genome named `GCF_000001405.34_GRCh38.p8_protein.faa.gz` is downloaded. The return value of `getProteome()` is the folder path to the downloaded proteome file
that can then be used as input to the `read_proteome()` function. The variable `HS.proteome.refseq` stores the path to the downloaded proteome. 
Subsequently, users can use the `read_proteome()` function to import the proteome into the R session. Users can choose to work with the proteome sequence in R either as [Biostrings](http://bioconductor.org/packages/release/bioc/html/Biostrings.html) object (`obj.type = "Biostrings"`) or [data.table](https://github.com/Rdatatable/data.table/wiki) object
(`obj.type = "data.table"`) by specifying the `obj.type` argument of the `read_proteome()` function. 


```{r,eval=FALSE}
# import proteome as Biostrings object
Human_Proteome <- read_proteome(file = HS.proteome.refseq, 
                                obj.type = "Biostrings")
```

```{r,eval=FALSE}
Human_Proteome
```

```
  A AAStringSet instance of length 109018
         width seq                                                 names               
     [1]  1474 MGKNKLLHPSLVLLLLVLLPTDAS...DYYETDEFAIAEYNAPCSKDLGNA NP_000005.2 alpha...
     [2]   290 MDIEAYFERIGYKNSRNKLDLETL...LRNIFKISLGRNLVPKPGDGSLTI NP_000006.2 aryla...
     [3]   421 MAAGFGRCCRVLRSISRFHWRSQH...QIYEGTSQIQRLIVAREHIDKYKN NP_000007.1 mediu...
     [4]   412 MAAALLARASGPARRALCPRAWRQ...EIYEGTSEIQRLVIAGHLLRSYRS NP_000008.1 short...
     [5]   655 MQAARMAASLGRQLLRLGGGSSRL...RNFKSISKALVERGGVVTSNPLGF NP_000009.1 very ...
     ...   ... ...
[109014]    98 MPLIYMNIMLAFTISLLGMLVYRS...LALLVSISNTYGLDYVHNLNLLQC YP_003024034.1 NA...
[109015]   459 MLKLIVPTIMLLPLTWLSKKHMIW...LMFMHLSPILLLSLNPDIITGFSS YP_003024035.1 NA...
[109016]   603 MTMHTTMTTLTLTSLIPPILTTLV...QKGMIKLYFLSFFFPLILTLLLIT YP_003024036.1 NA...
[109017]   174 MMYALFLLSVGLVMGFVGFSSKPS...WLVVVTGWTLFVGVYIVIEIARGN YP_003024037.1 NA...
[109018]   380 MTPMRKTNPLMKLINHSFIDLPTP...LYFTTILILMPTISLIENKMLKWA YP_003024038.1 cy...
```

Alternatively, users can perform the pipeline logic of the [magrittr](https://github.com/smbache/magrittr) package:

```{r,eval=FALSE}
# install.packages("magrittr")
library(magrittr)
# import proteome as Biostrings object
Human_Proteome <- getProteome( db       = "refseq", 
                               organism = "Homo sapiens",
                               path     = file.path("_ncbi_downloads","proteomes")) %>%
    read_proteome(obj.type = "Biostrings")
```

```{r,eval=FALSE}
Human_Proteome
```

```
  A AAStringSet instance of length 109018
         width seq                                                 names               
     [1]  1474 MGKNKLLHPSLVLLLLVLLPTDAS...DYYETDEFAIAEYNAPCSKDLGNA NP_000005.2 alpha...
     [2]   290 MDIEAYFERIGYKNSRNKLDLETL...LRNIFKISLGRNLVPKPGDGSLTI NP_000006.2 aryla...
     [3]   421 MAAGFGRCCRVLRSISRFHWRSQH...QIYEGTSQIQRLIVAREHIDKYKN NP_000007.1 mediu...
     [4]   412 MAAALLARASGPARRALCPRAWRQ...EIYEGTSEIQRLVIAGHLLRSYRS NP_000008.1 short...
     [5]   655 MQAARMAASLGRQLLRLGGGSSRL...RNFKSISKALVERGGVVTSNPLGF NP_000009.1 very ...
     ...   ... ...
[109014]    98 MPLIYMNIMLAFTISLLGMLVYRS...LALLVSISNTYGLDYVHNLNLLQC YP_003024034.1 NA...
[109015]   459 MLKLIVPTIMLLPLTWLSKKHMIW...LMFMHLSPILLLSLNPDIITGFSS YP_003024035.1 NA...
[109016]   603 MTMHTTMTTLTLTSLIPPILTTLV...QKGMIKLYFLSFFFPLILTLLLIT YP_003024036.1 NA...
[109017]   174 MMYALFLLSVGLVMGFVGFSSKPS...WLVVVTGWTLFVGVYIVIEIARGN YP_003024037.1 NA...
[109018]   380 MTPMRKTNPLMKLINHSFIDLPTP...LYFTTILILMPTISLIENKMLKWA YP_003024038.1 cy...
```

### Example `NCBI Genbank`:

```{r,eval=FALSE}
# download the proteome of Homo sapiens from genbank
# and store the corresponding proteome file in '_ncbi_downloads/proteomes'
HS.proteome.genbank <- getProteome( db       = "genbank", 
                                   organism = "Homo sapiens",
                                   path     = file.path("_ncbi_downloads","proteomes"))
```


```{r,eval=FALSE}
# import proteome as Biostrings object
Human_Proteome <- read_proteome(file = HS.proteome.genbank, 
                                obj.type = "Biostrings")
```

```{r,eval=FALSE}
Human_Proteome
```

```
  A AAStringSet instance of length 13
     width seq                                                     names               
 [1]   318 MPMANLLLLIVPILIAMAFLMLTERK...FLPLTLALLMWYVSMPITISSIPPQT AAB58943.1 NADH d...
 [2]   347 MNPLAQPVIYSTIFAGTLITALSSHW...PTPFLPTLIALTTLLLPISPFMLMIL AAB58944.1 NADH d...
 [3]   513 MFADRWLFSTNHKDIGTLYLLFGAWA...PSMNLEWLYGCPPPYHTFEEPVYMKS AAB58945.1 cytoch...
 [4]   227 MAHAAQVGLQDATSPIMEELITFHDH...ANHSFMPIVLELIPLKIFEMGPVFTL AAB58946.1 cytoch...
 [5]    68 MPQLNTTVWPTMITPMLLTLFLITQL...KMKNYNKPWEPKWTKICSLHSLPPQS AAB58947.1 ATPase...
 ...   ... ...
 [9]    98 MPLIYMNIMLAFTISLLGMLVYRSHL...VGLALLVSISNTYGLDYVHNLNLLQC AAB58951.1 NADH d...
[10]   459 MLKLIVPTIMLLPLTWLSKKHMIWIN...NTLMFMHLSPILLLSLNPDIITGFSS AAB58952.1 NADH d...
[11]   603 MTMHTTMTTLTLTSLIPPILTTLVNP...STQKGMIKLYFLSFFFPLILTLLLIT AAB58953.1 NADH d...
[12]   174 MMYALFLLSVGLVMGFVGFSSKPSPI...GRWLVVVTGWTLFVGVYIVIEIARGN AAB58954.1 NADH d...
[13]   380 MTPMRKTNPLMKLINHSFIDLPTPSN...SVLYFTTILILMPTISLIENKMLKWA AAB58955.3 cytoch...
```

### Example `ENSEMBL`:

```{r,eval=FALSE}
# download the proteome of Homo sapiens from ENSEMBL
# and store the corresponding proteome file in '_ncbi_downloads/proteomes'
HS.proteome.ensembl <- getProteome( db       = "ensembl", 
                                   organism = "Homo sapiens",
                                   path     = file.path("_ncbi_downloads","proteomes"))
```

```{r,eval=FALSE}
# import proteome as Biostrings object
Human_Proteome <- read_proteome(file = HS.proteome.ensembl, 
                                obj.type = "Biostrings")
```

```{r,eval=FALSE}
Human_Proteome
```

```
  A AAStringSet instance of length 102915
         width seq                                                 names               
     [1]     4 TGGY                                                ENSP00000452494.1...
     [2]     4 GTGG                                                ENSP00000488240.1...
     [3]     4 GTGG                                                ENSP00000487941.1...
     [4]     3 PSY                                                 ENSP00000451515.1...
     [5]     2 EI                                                  ENSP00000451042.1...
     ...   ... ...
[102911]   554 MLLPLLLSSLLGGSQAMDGRFWIR...GVRPRPEARMPKGTQADYAEVKFQ ENSP00000414324.2...
[102912]   639 MLLPLLLSSLLGGSQAMDGRFWIR...GVRPRPEARMPKGTQADYAEVKFQ ENSP00000389132.2...
[102913]   697 MLLPLLLSSLLGGSQAMDGRFWIR...GVRPRPEARMPKGTQADYAEVKFQ ENSP00000345243.4...
[102914]   155 MLLPLLLSSLLGVLSFTPRPQDHN...SGRYTCRAENRLGSQQRALDLSVQ ENSP00000435281.1...
[102915]   186 MRRCYCHCCCPRCWADWTGSTPAY...HFSVLSFTPRPQDHNTDLTCHVDF ENSP00000433838.1...
```


### Example `ENSEMBLGENOMES`:

Due to the unavailability of `"Homo sapiens"` at `db = "ensemblgenomes"` here we choose `"Arabidopsis thaliana"`
as example.

```{r,eval=FALSE}
# download the proteome of Arabidopsis thaliana from ENSEMBLGENOMES
# and store the corresponding proteome file in '_ncbi_downloads/proteomes'
AT.proteome.ensemblgenomes <- getProteome( db       = "ensemblgenomes", 
                                           organism = "Arabidopsis thaliana",
                                           path     = file.path("_ncbi_downloads","proteomes"))
```


```{r,eval=FALSE}
# import proteome as Biostrings object
Cress_Proteome <- read_proteome(file = AT.proteome.ensemblgenomes, 
                                obj.type = "Biostrings")
```

```{r,eval=FALSE}
Cress_Proteome
```

```
  A AAStringSet instance of length 35386
        width seq                                                  names               
    [1]   415 MGRDETETYITVPSFFKCPISLDVM...IKVLKFNSSALAAYETKTTHIMPF AT3G18710.1 pep:k...
    [2]   855 MATENPIRISGSNERWSNSRKVSVP...YTYGKHIVSRLEQPSIEENQELRR AT4G25880.2 pep:k...
    [3]   858 MATENPIRISGSNERWSNSRKVSVP...GKHIVSRLEQPSIEGMKFPNKTKN AT4G25880.3 pep:k...
    [4]   861 MATENPIRISGSNERWSNSRKVSVP...YTYGKHIVSRLEQPSIEENQELRR AT4G25880.1 pep:k...
    [5]   358 MTKAYSTRVLTFLILISLMAVTLNL...CSARNTQSFMSVLEEGIEEAISMI AT1G71695.1 pep:k...
    ...   ... ...
[35382]   374 MHSRSALLYRFLRPASRCFSSSSAV...YKAGEYYIKSMIEADRVASPSTSP AT2G20860.1 pep:k...
[35383]   392 MADNLNLVSVLGVLLVLTIFHNPII...WEPNNLAIRRRPSRDFYLGLAAYY AT3G14210.1 pep:k...
[35384]   495 MASLLSPATPTATSAAFHSCSTAGF...LRHPYFLLGGDQAAAVLSKLSFSK AT5G01920.1 pep:k...
[35385]   563 MSLTKKASEPKLSGTSIKPTTLNPH...VAVQRYLLEEEGLDYSEPQAGLLR AT2G26280.1 pep:k...
[35386]   453 MGVSLLKQQHRITNQADTFSRFMER...NHQQQQQQQRSELRVENGLANNVI AT4G32600.1 pep:k...
```

### Example `Uniprot`:

Another  way of retrieving proteome sequences is from the [UniProt](http://www.uniprot.org/) database.

```{r,eval=FALSE}
# download the proteome of Mus musculus from UniProt
# and store the corresponding proteome file in '_ncbi_downloads/proteomes'
Mm.proteome.uniprot<- getProteome( db       = "uniprot", 
                                   organism = "Mus musculus",
                                   path     = file.path("_ncbi_downloads","proteomes"))
```


```{r,eval=FALSE}
# import proteome as Biostrings object
Mouse_Proteome <- read_proteome(file = Mm.proteome.uniprot)
```

```{r,eval=FALSE}
Mouse_Proteome
```

```
A AAStringSet instance of length 84522
        width seq                                  names               
    [1]   781 MATQADLMELDMAMEPD...LPPGDSNQLAWFDTDL sp|Q02248|CTNB1_M...
    [2]  2531 MPRLLTPLLCLTLLPAL...PTTMPSQITHIPEAFK sp|Q01705|NOTC1_M...
    [3]   437 MLLLLARCFLVILASSL...LDSETMHPLGMAVKSS sp|Q62226|SHH_MOU...
    [4]   380 MKKPIGILSPGVALGTA...VKCKKCTEIVDQFVCK sp|P22725|WNT5A_M...
    [5]   387 MEESQSDISLELPLSQE...SRHKKTMVKKVGPDSD sp|P02340|P53_MOU...
    ...   ... ...
[84518]   459 MLKIILPSLMLLPLTWL...LILLTTSPKLITGLTM tr|A0A0F6PZ84|A0A...
[84519]   380 MTNMRKTHPLFKIINHS...LMPISGIIEDKMLKLY tr|U3TEV9|U3TEV9_...
[84520]   381 MTNMRKTHPLFKIINHS...MPISGIIEDKMLKLYP tr|A0A0F6PXN8|A0A...
[84521]   172 MNNYIFVLSSLFLVGCL...WSLFAGIFIIIEITRD tr|A0A0F6PXK9|A0A...
[84522]   381 MTNMRKTHPLFKIINHS...MPISGIIEDKMLKLYP tr|A0A0F6PYR5|A0A...
```


### CDS Retrieval

The `getCDS()` function is an interface function to the [NCBI RefSeq](http://www.ncbi.nlm.nih.gov/refseq/about/), [NCBI Genbank](http://www.ncbi.nlm.nih.gov/genbank/about/),
[ENSEMBL](http://www.ensembl.org/index.html) and [ENSEMBLGENOMES](http://ensemblgenomes.org/) databases from
which corresponding CDS files can be retrieved. It works analogous to `getGenome()` and `getProteome()`. 

The `db` argument specifies from which database proteomes in `*.fasta` file format shall be retrieved.

Options are:

- `db = "refseq"` for retrieval from [NCBI RefSeq](http://www.ncbi.nlm.nih.gov/refseq/about/)
- `db = "genbank"` for retrieval from [NCBI Genbank](http://www.ncbi.nlm.nih.gov/genbank/about/)
- `db = "ensembl"` for retrieval from [ENSEMBL](http://www.ensembl.org/index.html)
- `db = "ensemblgenomes"` for retrieval from [ENSEMBLGENOMES](http://ensemblgenomes.org/)

Furthermore, again users need to specify the scientific name of the organism of interest for which a proteomes shall be downloaded, e.g. `organism = "Homo sapiens"`.
Finally, the `path` argument specifies the folder path in which the corresponding CDS file shall be locally stored. In case users would like to store the CDS file at a different location,
they can specify the `path = file.path("put","your","path","here")` argument.

### Example `NCBI RefSeq`:

```{r,eval=FALSE}
# download the genome of Homo sapiens from refseq
# and store the corresponding genome CDS file in '_ncbi_downloads/CDS'
HS.cds.refseq <- getCDS( db       = "refseq", 
                         organism = "Homo sapiens",
                         path     = file.path("_ncbi_downloads","CDS"))
```


In this example, `getCDS()` creates a directory named `'_ncbi_downloads/CDS'` into which the corresponding genome named `Homo_sapiens_cds_from_genomic_refseq.fna.gz` is downloaded. The return value of `getCDS()` is the folder path to the downloaded genome file
that can then be used as input to the `read_cds()` function. The variable `HS.cds.refseq` stores the path to the downloaded CDS file. 
Subsequently, users can use the `read_cds()` function to import the genome into the R session. Users can choose to work with the genome sequence in R either as [Biostrings](http://bioconductor.org/packages/release/bioc/html/Biostrings.html) object (`obj.type = "Biostrings"`) or [data.table](https://github.com/Rdatatable/data.table/wiki) object
(`obj.type = "data.table"`) by specifying the `obj.type` argument of the `read_cds()` function. 

```{r,eval=FALSE}
# import downloaded CDS as Biostrings object
Human_CDS <- read_cds(file     = HS.cds.refseq, 
                      obj.type = "Biostrings")

```

```{r,eval=FALSE}
# look at the Biostrings object
Human_CDS
```

```
 A BStringSet instance of length 114967
          width seq                                                names               
     [1]    918 ATGGTGACTGAATTCATTTTTCTG...CACATTCTAGTGTAAAGTTTTAG lcl|NC_000001.11_...
     [2]    402 ATGAGTGACAGCATCAACTTCTCT...CAGGACCCAGGCACAGGCATTAG lcl|NC_000001.11_...
     [3]    402 ATGAGTGACAGCATCAACTTCTCT...CAGGACCCAGGCACAGGCATTAG lcl|NC_000001.11_...
     [4]    402 ATGAGTGACAGCATCAACTTCTCT...CAGGACCCAGGCACAGGCATTAG lcl|NC_000001.11_...
     [5]    402 ATGAGTGACAGCATCAACTTCTCT...CAGGACCCAGGCACAGGCATTAG lcl|NC_000001.11_...
     ...    ... ...
[114963]    297 ATGCCCCTCATTTACATAAATATT...ACCTAAACCTACTCCAATGCTAA lcl|NC_012920.1_c...
[114964]   1378 ATGCTAAAACTAATCGTCCCAACA...CATCATTACCGGGTTTTCCTCTT lcl|NC_012920.1_c...
[114965]   1812 ATAACCATGCACACTACTATAACC...TAACCCTACTCCTAATCACATAA lcl|NC_012920.1_c...
[114966]    525 ATGATGTATGCTTTGTTTCTGTTG...TTGAGATTGCTCGGGGGAATAGG lcl|NC_012920.1_c...
[114967]   1141 ATGACCCCAATACGCAAAACTAAC...AAACAAAATACTCAAATGGGCCT lcl|NC_012920.1_c...
```

Internally, a text file named `doc_Homo_sapiens_db_refseq.txt` is generated. The information stored in this log file is structured as follows:

```
File Name: Homo_sapiens_cds_from_genomic_refseq.fna.gz
Organism Name: Homo_sapiens
Database: NCBI refseq
URL: ftp://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/001/405/
GCF_000001405.35_GRCh38.p9/GCF_000001405.35_GRCh38.p9_cds_from_genomic.fna.gz
Download_Date: Sun Oct 23 17:19:05 2016
refseq_category: reference genome
assembly_accession: GCF_000001405.35
bioproject: PRJNA168
biosample: NA
taxid: 9606
infraspecific_name: NA
version_status: latest
release_type: Patch
genome_rep: Full
seq_rel_date: 2016-09-26
submitter: Genome Reference Consortium
```

In summary, the `getCDS()` and `read_cds()` functions allow users to retrieve CDS files by specifying
the scientific name of the organism of interest and allow them to import the retrieved CDS file e.g. as `Biostrings` object.
Thus, users can then perform the `Biostrings notation` to work with downloaded CDS and can rely on the log
file generated by `getCDS()` to better document the source and version of CDS used for subsequent studies.

Alternatively, users can perform the pipeline logic of the [magrittr](https://github.com/smbache/magrittr) package:

```{r,eval=FALSE}
# install.packages("magrittr")
library(magrittr)
# import CDS as Biostrings object
Human_CDS <- getCDS( db       = "refseq", 
                     organism = "Homo sapiens",
                     path     = file.path("_ncbi_downloads","CDS")) %>%
    read_cds(obj.type = "Biostrings")
```

```{r,eval=FALSE}
Human_CDS
```

```
 A BStringSet instance of length 114967
          width seq                                                names               
     [1]    918 ATGGTGACTGAATTCATTTTTCTG...CACATTCTAGTGTAAAGTTTTAG lcl|NC_000001.11_...
     [2]    402 ATGAGTGACAGCATCAACTTCTCT...CAGGACCCAGGCACAGGCATTAG lcl|NC_000001.11_...
     [3]    402 ATGAGTGACAGCATCAACTTCTCT...CAGGACCCAGGCACAGGCATTAG lcl|NC_000001.11_...
     [4]    402 ATGAGTGACAGCATCAACTTCTCT...CAGGACCCAGGCACAGGCATTAG lcl|NC_000001.11_...
     [5]    402 ATGAGTGACAGCATCAACTTCTCT...CAGGACCCAGGCACAGGCATTAG lcl|NC_000001.11_...
     ...    ... ...
[114963]    297 ATGCCCCTCATTTACATAAATATT...ACCTAAACCTACTCCAATGCTAA lcl|NC_012920.1_c...
[114964]   1378 ATGCTAAAACTAATCGTCCCAACA...CATCATTACCGGGTTTTCCTCTT lcl|NC_012920.1_c...
[114965]   1812 ATAACCATGCACACTACTATAACC...TAACCCTACTCCTAATCACATAA lcl|NC_012920.1_c...
[114966]    525 ATGATGTATGCTTTGTTTCTGTTG...TTGAGATTGCTCGGGGGAATAGG lcl|NC_012920.1_c...
[114967]   1141 ATGACCCCAATACGCAAAACTAAC...AAACAAAATACTCAAATGGGCCT lcl|NC_012920.1_c...
```

### Example `NCBI Genbank`:

```{r,eval=FALSE}
# download the genome of Homo sapiens from genbank
# and store the corresponding genome CDS file in '_ncbi_downloads/CDS'
HS.cds.genbank <- getCDS( db       = "genbank", 
                         organism = "Homo sapiens",
                         path     = file.path("_ncbi_downloads","CDS"))
```

```{r,eval=FALSE}
# import downloaded CDS as Biostrings object
Human_CDS <- read_cds(file     = HS.cds.genbank, 
                      obj.type = "Biostrings")

```

```{r,eval=FALSE}
# look at the Biostrings object
Human_CDS
```

```
  A BStringSet instance of length 13
     width seq                                                     names               
 [1]   956 ATACCCATGGCCAACCTCCTACTCCT...ATCTCCAGCATTCCCCCTCAAACCTA lcl|J01415.2_cds_...
 [2]  1042 ATTAATCCCCTGGCCCAACCCGTCAT...CTCCCCTTTTATACTAATAATCTTAT lcl|J01415.2_cds_...
 [3]  1542 ATGTTCGCCGACCGTTGACTATTCTC...AAGAACCCGTATACATAAAATCTAGA lcl|J01415.2_cds_...
 [4]   684 ATGGCACATGCAGCGCAAGTAGGTCT...AAATAGGGCCCGTATTTACCCTATAG lcl|J01415.2_cds_...
 [5]   207 ATGCCCCAACTAAATACTACCGTATG...TTCATTCATTGCCCCCACAATCCTAG lcl|J01415.2_cds_...
 ...   ... ...
 [9]   297 ATGCCCCTCATTTACATAAATATTAT...ATAACCTAAACCTACTCCAATGCTAA lcl|J01415.2_cds_...
[10]  1378 ATGCTAAAACTAATCGTCCCAACAAT...CGACATCATTACCGGGTTTTCCTCTT lcl|J01415.2_cds_...
[11]  1812 ATAACCATGCACACTACTATAACCAC...TCCTAACCCTACTCCTAATCACATAA lcl|J01415.2_cds_...
[12]   525 ATGATGTATGCTTTGTTTCTGTTGAG...TAATTGAGATTGCTCGGGGGAATAGG lcl|J01415.2_cds_...
[13]  1141 ATGACCCCAATACGCAAAACTAACCC...TGAAAACAAAATACTCAAATGGGCCT lcl|J01415.2_cds_...
```

### Example `ENSEMBL`:

```{r,eval=FALSE}
# download the genome of Homo sapiens from ensembl
# and store the corresponding genome CDS file in '_ncbi_downloads/CDS'
HS.cds.ensembl <- getCDS( db       = "ensembl", 
                         organism = "Homo sapiens",
                         path     = file.path("_ncbi_downloads","CDS"))
```

```{r,eval=FALSE}
# import downloaded CDS as Biostrings object
Human_CDS <- read_cds(file     = HS.cds.ensembl, 
                      obj.type = "Biostrings")

```

```{r,eval=FALSE}
# look at the Biostrings object
Human_CDS
```

```
  A BStringSet instance of length 102915
          width seq                                                names               
     [1]     13 ACTGGGGGATACG                                      ENST00000448914.1...
     [2]     12 GGGACAGGGGGC                                       ENST00000631435.1...
     [3]     12 GGGACAGGGGGC                                       ENST00000632684.1...
     [4]      9 CCTTCCTAC                                          ENST00000434970.2...
     [5]      8 GAAATAGT                                           ENST00000415118.1...
     ...    ... ...
[102911]   1665 ATGCTACTGCCACTGCTGCTGTCC...ATGCAGAAGTCAAGTTCCAATGA ENST00000436984.6...
[102912]   1920 ATGCTACTGCCACTGCTGCTGTCC...ATGCAGAAGTCAAGTTCCAATGA ENST00000439889.6...
[102913]   2094 ATGCTACTGCCACTGCTGCTGTCC...ATGCAGAAGTCAAGTTCCAATGA ENST00000339313.9...
[102914]    466 ATGCTACTGCCACTGCTGCTGTCC...AGCCCTGGACCTCTCTGTGCAGT ENST00000529627.1...
[102915]    559 ATGCGGAGATGCTACTGCCACTGC...CCTCACCTGCCATGTGGACTTCT ENST00000530476.1...
```

### Example `ENSEMBLGENOMES`:

Due to the inavailability of `"Homo sapiens"` at `db = "ensemblgenomes"` here we choose `"Arabidopsis thaliana"`
as example.

```{r,eval=FALSE}
# download the genome of Homo sapiens from ensemblgenomes
# and store the corresponding genome CDS file in '_ncbi_downloads/CDS'
AT.cds.ensemblgenomes <- getCDS( db       = "ensemblgenomes", 
                                 organism = "Arabidopsis thaliana",
                                 path     = file.path("_ncbi_downloads","CDS"))
```

```{r,eval=FALSE}
# import downloaded CDS as Biostrings object
Cress_CDS <- read_cds(file     = AT.cds.ensemblgenomes, 
                      obj.type = "Biostrings")

```

```{r,eval=FALSE}
# look at the Biostrings object
Cress_CDS
```

```
  A BStringSet instance of length 35386
        width seq                                                  names               
    [1]  1248 ATGGGGAGAGATGAAACAGAGACGT...ACTACTCATATTATGCCTTTTTGA AT3G18710.1 cds:k...
    [2]  2568 ATGGCAACTGAGAATCCTATTAGGA...GAAAACCAAGAATTGAGGAGATGA AT4G25880.2 cds:k...
    [3]  2577 ATGGCAACTGAGAATCCTATTAGGA...TTCCCCAATAAAACCAAGAATTGA AT4G25880.3 cds:k...
    [4]  2586 ATGGCAACTGAGAATCCTATTAGGA...GAAAACCAAGAATTGAGGAGATGA AT4G25880.1 cds:k...
    [5]  1077 ATGACAAAGGCTTATTCAACACGTG...GAGGAAGCTATTTCCATGATCTAA AT1G71695.1 cds:k...
    ...   ... ...
[35382]  1125 ATGCATTCGCGCTCCGCCTTGCTCT...GCTTCTCCTTCTACATCCCCGTAG AT2G20860.1 cds:k...
[35383]  1179 ATGGCAGACAATTTGAATTTGGTGA...TTGGGCCTCGCCGCCTATTATTAG AT3G14210.1 cds:k...
[35384]  1488 ATGGCCTCTCTTCTCTCTCCCGCCA...TCAAAGCTCAGTTTCAGCAAGTGA AT5G01920.1 cds:k...
[35385]  1689 ATGAGTTTAACAAAGAAAGCAAGTG...GAACCACAGGCCGGTCTCCTTAGA AT2G26280.1 cds:k...
[35386]  1362 ATGGGTGTTTCTCTTCTGAAACAAC...GGTTTGGCCAATAATGTTATCTAG AT4G32600.1 cds:k...
```


### RNA Retrieval

The `getRNA()` function is an interface function to the [NCBI RefSeq](http://www.ncbi.nlm.nih.gov/refseq/about/), [NCBI Genbank](http://www.ncbi.nlm.nih.gov/genbank/about/),
[ENSEMBL](http://www.ensembl.org/index.html) and [ENSEMBLGENOMES](http://ensemblgenomes.org/) databases from
which corresponding RNA files can be retrieved. It works analogous to `getGenome()`, `getProteome()`, and `getCDS()`. 

The `db` argument specifies from which database proteomes in `*.fasta` file format shall be retrieved.

Options are:

- `db = "refseq"` for retrieval from [NCBI RefSeq](http://www.ncbi.nlm.nih.gov/refseq/about/)
- `db = "genbank"` for retrieval from [NCBI Genbank](http://www.ncbi.nlm.nih.gov/genbank/about/)
- `db = "ensembl"` for retrieval from [ENSEMBL](http://www.ensembl.org/index.html)
- `db = "ensemblgenomes"` for retrieval from [ENSEMBLGENOMES](http://ensemblgenomes.org/)

Furthermore, again users need to specify the scientific name of the organism of interest for which a proteomes shall be downloaded, e.g. `organism = "Homo sapiens"`.
Finally, the `path` argument specifies the folder path in which the corresponding RNA file shall be locally stored. In case users would like to store the RNA file at a different location,
they can specify the `path = file.path("put","your","path","here")` argument.

### Example `NCBI RefSeq`:

```{r,eval=FALSE}
# download the RNA of Homo sapiens from refseq
# and store the corresponding RNA file in '_ncbi_downloads/RNA'
HS.rna.refseq <- getRNA( db       = "refseq", 
                         organism = "Homo sapiens",
                         path     = file.path("_ncbi_downloads","RNA"))
```


In this example, `getRNA()` creates a directory named `'_ncbi_downloads/RNA'` into which the corresponding RNA file named `Homo_sapiens_rna_from_genomic_refseq.fna.gz` is downloaded. The return value of `getRNA()` is the folder path to the downloaded genome file
that can then be used as input to the `read_rna()` function. The variable `HS.rna.refseq` stores the path to the downloaded RNA file. 
Subsequently, users can use the `read_cds()` function to import the genome into the R session. Users can choose to work with the genome sequence in R either as [Biostrings](http://bioconductor.org/packages/release/bioc/html/Biostrings.html) object (`obj.type = "Biostrings"`) or [data.table](https://github.com/Rdatatable/data.table/wiki) object
(`obj.type = "data.table"`) by specifying the `obj.type` argument of the `read_rna()` function. 

```{r,eval=FALSE}
# import downloaded RNA as Biostrings object
Human_rna <- read_rna(file     = HS.rna.refseq, 
                      obj.type = "Biostrings")

```

```{r,eval=FALSE}
# look at the Biostrings object
Human_rna
```

```
   A BStringSet instance of length 164136
          width seq                                                                                       names               
     [1]   1652 CTTGCCGTCAGCCTTTTCTTTGACCTCTTCTTTCTGTTCATGT...CACAGCTAGAGATCCTTTATTAAAAGCACACTGTTGGTTTCTG lcl|NC_000001.11_...
     [2]   1769 TCCGGCAGAGCGGAAGCGGCGGCGGGAGCTTCCGGGAGGGCGG...ACCAACAGTGTGCTTTTAATAAAGGATCTCTAGCTGTGCAGGA lcl|NC_000001.11_...
     [3]     68 TGTGGGAGAGGAACATGGGCTCAGGACAGCGGGTGTCAGCTTGCCTGACCCCCATGTCGCCTCTGTAG                      lcl|NC_000001.11_...
     [4]     23 TGACCCCCATGTCGCCTCTGTAG                                                                   lcl|NC_000001.11_...
     [5]     23 GAGAGGAACATGGGCTCAGGACA                                                                   lcl|NC_000001.11_...
     ...    ... ...
[164132]     59 GAGAAAGCTCACAAGAACTGCTAACTCATGCCCCCATGTCTAACAACATGGCTTTCTCA                               lcl|NC_012920.1_t...
[164133]     71 ACTTTTAAAGGATAACAGCTATCCATTGGTCTTAGGCCCCAAAAATTTTGGTGCAACTCCAAATAAAAGTA                   lcl|NC_012920.1_t...
[164134]     69 GTTCTTGTAGTTGAAATACAACGATGGTTTTTCATATCATTGGTCGTGGTTGTAGTCCGTGCGAGAATA                     lcl|NC_012920.1_t...
[164135]     66 GTCCTTGTAGTATAAACTAATACACCAGTCTTGTAAACCGGAGATGAAAACCTTTTTCCAAGGACA                        lcl|NC_012920.1_t...
[164136]     68 CAGAGAATAGTTTAAATTAGAATCTTAGCTTTGGGTGCTAATGGTGGAGTTAAAGACTTTTTCTCTGA                      lcl|NC_012920.1_t...
```

Internally, a text file named `doc_Homo_sapiens_db_refseq.txt` is generated. The information stored in this log file is structured as follows:

```
File Name: Homo_sapiens_rna_from_genomic_refseq.fna.gz
Organism Name: Homo_sapiens
Database: NCBI refseq
URL: ftp://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/001/405/GCF_000001405.36_GRCh38.p10/GCF_000001405.36_GRCh38.p10_rna_from_genomic.fna.gz
Download_Date: Wed Mar 15 16:46:45 2017
refseq_category: reference genome
assembly_accession: GCF_000001405.36
bioproject: PRJNA168
biosample: NA
taxid: 9606
infraspecific_name: NA
version_status: latest
release_type: Patch
genome_rep: Full
seq_rel_date: 2017-01-06
submitter: Genome Reference Consortium
```

In summary, the `getRNA()` and `read_rna()` functions allow users to retrieve RNA files by specifying
the scientific name of the organism of interest and allow them to import the retrieved RNA file e.g. as `Biostrings` object.
Thus, users can then perform the `Biostrings notation` to work with downloaded RNA and can rely on the log
file generated by `getRNA()` to better document the source and version of RNA used for subsequent studies.

Alternatively, users can perform the pipeline logic of the [magrittr](https://github.com/smbache/magrittr) package:

```{r,eval=FALSE}
# install.packages("magrittr")
library(magrittr)
# import RNA as Biostrings object
Human_rna <- getRNA( db       = "refseq", 
                     organism = "Homo sapiens",
                     path     = file.path("_ncbi_downloads","RNA")) %>%
    read_cds(obj.type = "Biostrings")
```

```{r,eval=FALSE}
Human_rna
```

```
   A BStringSet instance of length 164136
          width seq                                                                                       names               
     [1]   1652 CTTGCCGTCAGCCTTTTCTTTGACCTCTTCTTTCTGTTCATGT...CACAGCTAGAGATCCTTTATTAAAAGCACACTGTTGGTTTCTG lcl|NC_000001.11_...
     [2]   1769 TCCGGCAGAGCGGAAGCGGCGGCGGGAGCTTCCGGGAGGGCGG...ACCAACAGTGTGCTTTTAATAAAGGATCTCTAGCTGTGCAGGA lcl|NC_000001.11_...
     [3]     68 TGTGGGAGAGGAACATGGGCTCAGGACAGCGGGTGTCAGCTTGCCTGACCCCCATGTCGCCTCTGTAG                      lcl|NC_000001.11_...
     [4]     23 TGACCCCCATGTCGCCTCTGTAG                                                                   lcl|NC_000001.11_...
     [5]     23 GAGAGGAACATGGGCTCAGGACA                                                                   lcl|NC_000001.11_...
     ...    ... ...
[164132]     59 GAGAAAGCTCACAAGAACTGCTAACTCATGCCCCCATGTCTAACAACATGGCTTTCTCA                               lcl|NC_012920.1_t...
[164133]     71 ACTTTTAAAGGATAACAGCTATCCATTGGTCTTAGGCCCCAAAAATTTTGGTGCAACTCCAAATAAAAGTA                   lcl|NC_012920.1_t...
[164134]     69 GTTCTTGTAGTTGAAATACAACGATGGTTTTTCATATCATTGGTCGTGGTTGTAGTCCGTGCGAGAATA                     lcl|NC_012920.1_t...
[164135]     66 GTCCTTGTAGTATAAACTAATACACCAGTCTTGTAAACCGGAGATGAAAACCTTTTTCCAAGGACA                        lcl|NC_012920.1_t...
[164136]     68 CAGAGAATAGTTTAAATTAGAATCTTAGCTTTGGGTGCTAATGGTGGAGTTAAAGACTTTTTCTCTGA                      lcl|NC_012920.1_t...
```

### Example `NCBI Genbank`:

```{r,eval=FALSE}
# download the RNA of Homo sapiens from genbank
# and store the corresponding genome RNA file in '_ncbi_downloads/RNA'
HS.rna.genbank <- getRNA( db       = "genbank", 
                         organism = "Homo sapiens",
                         path     = file.path("_ncbi_downloads","RNA"))
```

```{r,eval=FALSE}
# import downloaded RNA as Biostrings object
Human_rna <- read_cds(file     = HS.rna.genbank, 
                      obj.type = "Biostrings")

```

```{r,eval=FALSE}
# look at the Biostrings object
Human_rna
```

```
  A BStringSet instance of length 24
     width seq                                                                                            names               
 [1]    71 GTTTATGTAGCTTACCTCCTCAAAGCAATACACTGAAAATGTTTAGACGGGCTCACATCACCCCATAAACA                        lcl|J01415.2_trna...
 [2]   954 AATAGGTTTGGTCCTAGCCTTTCTATTAGCTCTTAGTAAGATTACA...AAGTCGTAACATGGTAAGTGTACTGGAAAGTGCACTTGGACGAAC lcl|J01415.2_rrna...
 [3]    69 CAGAGTGTAGCTTAACACAAAGCACCCAACTTACACTTAGGAGATTTCAACTTAACTTGACCGCTCTGA                          lcl|J01415.2_trna...
 [4]  1559 GCTAAACCTAGCCCCAAACCCACTCCACCTTACTACCAGACAACCT...ATCTCAACTTAGTATTATACCCACACCCACCCAAGAACAGGGTTT lcl|J01415.2_rrna...
 [5]    75 GTTAAGATGGCAGAGCCCGGTAATCGCATAAAACTTAAAACTTTACAGTCAGAGGTTCAATTCCTCTTCTTAACA                    lcl|J01415.2_trna...
 ...   ... ...
[20]    59 GAGAAAGCTCACAAGAACTGCTAACTCATGCCCCCATGTCTAACAACATGGCTTTCTCA                                    lcl|J01415.2_trna...
[21]    71 ACTTTTAAAGGATAACAGCTATCCATTGGTCTTAGGCCCCAAAAATTTTGGTGCAACTCCAAATAAAAGTA                        lcl|J01415.2_trna...
[22]    69 GTTCTTGTAGTTGAAATACAACGATGGTTTTTCATATCATTGGTCGTGGTTGTAGTCCGTGCGAGAATA                          lcl|J01415.2_trna...
[23]    66 GTCCTTGTAGTATAAACTAATACACCAGTCTTGTAAACCGGAGATGAAAACCTTTTTCCAAGGACA                             lcl|J01415.2_trna...
[24]    68 CAGAGAATAGTTTAAATTAGAATCTTAGCTTTGGGTGCTAATGGTGGAGTTAAAGACTTTTTCTCTGA                           lcl|J01415.2_trna...
```

### Example `ENSEMBL`:

```{r,eval=FALSE}
# download the RNA of Homo sapiens from ensembl
# and store the corresponding genome RNA file in '_ncbi_downloads/RNA'
HS.rna.ensembl <- getRNA( db       = "ensembl", 
                          organism = "Homo sapiens",
                          path     = file.path("_ncbi_downloads","RNA"))
```

```{r,eval=FALSE}
# import downloaded RNA as Biostrings object
Human_rna <- read_cds(file     = HS.rna.ensembl, 
                      obj.type = "Biostrings")

```

```{r,eval=FALSE}
# look at the Biostrings object
Human_rna
```

```
  A BStringSet instance of length 36701
         width seq                                                                                        names               
    [1]    104 GTGCTCACTTTGGCAACATACATACTAAAATTGGACGGATACAG...GCACAAGGATGACATGCAAATTCATGAAGCATTCCATATTTTT ENST00000516494.2...
    [2]    164 TTAACTACCTGACAGAGGAGATACTGTGATCATGAAAGTGGTTT...CATAATTTCTGGTGGTAGGGGACTGCGTTCATGTTCTCCCCTA ENST00000627793.1...
    [3]    114 ACACTGGTTTCTCTTCAGATCGAATAAATCTTTCGCCTTTTACT...AGTTATAAGCTAATTTTTTGTAAGCCTTGCCCTGGGGAGGCAG ENST00000629478.1...
    [4]     64 CAGTGTTACACCTCTTTTAGAATTTATCTATCAGGTTTTCCAGTGTTCACTGAAATTTGTCTCT                           ENST00000629187.1...
    [5]    107 GTGTTGGCCTGGGCAGCACGTATACTAAAGTTGGAATGACACAG...GCGCAAGGATGATGTGCAAATTCGTGACAAGTTCCATATTTTT ENST00000631612.1...
    ...    ... ...
[36697]     90 GGCTAGTCCAAATGTAGTGGTGTTCCAAACTAATTAATCACAACCAGTTACAGATTTCTTGTTTTCTTTTCCACTCACACTTAGCCTTAA ENST00000410951.1...
[36698]    109 GGCTGATCTGAAGATGATGAGTTATCTCAATTGATTGTTCAGCC...TCTATTCTTTCCTCTCTTCTCACTACTGCACTTGGCTAGGAAA ENST00000410462.2...
[36699]    109 GGTGGGTCCAAAGGTAGCAAGTTATCTCAATTGATCACAGTCAG...CATTCTATCACCCCTTCTCATTACTGTACTTGACTAGTCTTTT ENST00000364501.1...
[36700]    293 GGATATGAGGGCGATCTGGCAGGGACATCTGTCACCCCACTGAT...AAAATTAGCTGGGCATAGTGGCGTGCACCTGTCGTCCTAGCTA ENST00000365097.1...
[36701]    172 TTGAAGGCGTGGAGACTGAAGTCCTCTCTATATCCACAGAACAG...AAGAGGGCTGTTCAGTCTCCATGCCCTTCAATCCTTGGCTACT ENST00000615087.1...
```

### Example `ENSEMBLGENOMES`:

Due to the inavailability of `"Homo sapiens"` at `db = "ensemblgenomes"` here we choose `"Arabidopsis thaliana"`
as example.

```{r,eval=FALSE}
# download the genome of Arabidopsis thaliana from ensemblgenomes
# and store the corresponding genome RNA file in '_ncbi_downloads/RNA'
AT.rna.ensemblgenomes <- getRNA( db       = "ensemblgenomes", 
                                 organism = "Arabidopsis thaliana",
                                 path     = file.path("_ncbi_downloads","RNA"))
```

```{r,eval=FALSE}
# import downloaded RNA as Biostrings object
Cress_rna <- read_cds(file     = AT.rna.ensemblgenomes, 
                      obj.type = "Biostrings")
```

```{r,eval=FALSE}
# look at the Biostrings object
Cress_rna
```

```
  A BStringSet instance of length 1398
       width seq                                                                                          names               
   [1]   101 TAGTGAGCACAAAGCGAACTATTCTTTCGCCTTTTACTAAAGAAT...CCACGCTAAGTGGCATACGCCTATTTTTGGAGGGATCTGCCTTC AT1G05853.1 ncrna...
   [2]   195 ATACCTTTCTCGGCCTTTTGGCTAAGATCAAGTGTAGTATCTGTT...AGTCGCCCATGCGTTGCACTACTGCATGGGCCTGGCTCAACCCG AT2G04455.1 ncrna...
   [3]   195 TTTTTCTAATTACTTGCAAGCGATGTGTAAATTAGATTAATTCAA...TCATAAATTTAGATAATATTTCACTAAATAAGTTGAAATTGACC AT5G09945.1 ncrna...
   [4]   101 CGCAGCTATGTGGTGAGCACAAAGCGAACTATTCTTTCGCCTTTT...CGTGTGCTCTCCACGCCAAGTGGCTTACGCCTATTTTTGGAGGG AT4G04185.1 ncrna...
   [5]    55 GTGATCAATAAGACAAGTGGCCTAGGCTAGTGATCGCATTGCACATATCGTTTGG                                      AT4G06405.1 ncrna...
   ...   ... ...
[1394]   118 ACACATAAAACTAGTATATTGCCTTGAACATGGTTTATTAGGAAA...TCCTTATATACAATGTTCAAAGTAATGAATTAGTTTTATTTTGT at4g21362 ncrna c...
[1395]   121 CTTACAGAGATCTTTGGCATTCTGTCCACCTCCTCTCTCTATATT...TTTCGTAAGAGGAGGTGGGCATACTGCCAATAGAGATCTGTTAG at1g76135 ncrna c...
[1396]   100 GGTTGGATTACTGGGCGAATACTCCTATGGCAGATCGCATTGGCT...TAAAATGCTTCTCTGCCAAAGGAGATTTGCCCCGCAATTCATCC at2g34202 ncrna c...
[1397]   174 TTGACCCAGCCAAATCGACTCACCGTTAAAACTTCTTAACAGCTC...CGCTGTTAAGGAGTGTTAACGGTGAGTCGATTTGGCCGGGTCAA at4g07565 ncrna c...
[1398]   189 CTTAAGGCAATTTGACCTATATCAAATTTGGCACCGTTGCCGGGG...CCCGGCAACGGCGCCAAATTTGATATAGCTCAAATCGCCTTAAG at2g05455 ncrna c...
```


### Retrieve the annotation file of a particular genome

Finally, users can download the corresponding annotation `.gff` files for particular genomes of interest using the `getGFF()` or alternatively for `ensembl` and `ensemblgenomes` the `getGTF()` function.

### Example `NCBI RefSeq`:

```{r,eval=FALSE}
# download the GFF file of Homo sapiens from refseq
# and store the corresponding file in '_ncbi_downloads/annotation'
HS.gff.refseq <- getGFF( db       = "refseq", 
                         organism = "Homo sapiens", 
                         path = file.path("_ncbi_downloads","annotation"))
```

After downloading the `.gff` file, users can import the `.gff` file with `read_gff()`.

```{r,eval=FALSE}
# import downloaded GFF file
Human_GFF <- read_gff(file = HS.gff.refseq)
```

```{r,eval=FALSE}
Human_GFF
```

```
          seqid     source       type start       end score strand phase
          <chr>      <chr>      <chr> <int>     <int> <dbl>  <chr> <dbl>
1  NC_000001.11     RefSeq     region     1 248956422     0      +     0
2  NC_000001.11 BestRefSeq       gene 11874     14409     0      +     0
3  NC_000001.11 BestRefSeq transcript 11874     14409     0      +     0
4  NC_000001.11 BestRefSeq       exon 11874     12227     0      +     0
5  NC_000001.11 BestRefSeq       exon 12613     12721     0      +     0
6  NC_000001.11 BestRefSeq       exon 13221     14409     0      +     0
7  NC_000001.11 BestRefSeq       gene 14362     29370     0      -     0
8  NC_000001.11 BestRefSeq transcript 14362     29370     0      -     0
9  NC_000001.11 BestRefSeq       exon 29321     29370     0      -     0
10 NC_000001.11 BestRefSeq       exon 24738     24891     0      -     0
```

### Example `NCBI Genbank`:

```{r,eval=FALSE}
# download the GFF file of Homo sapiens from genbank
# and store the corresponding file in '_ncbi_downloads/annotation'
HS.gff.genbank <- getGFF( db       = "genbank", 
                         organism = "Homo sapiens", 
                         path = file.path("_ncbi_downloads","annotation"))
```

After downloading the `.gff` file, users can import the `.gff` file with `read_gff()`.

```{r,eval=FALSE}
# import downloaded GFF file
Human_GFF <- read_gff(file = HS.gff.genbank)
```

```{r,eval=FALSE}
# show all elements of the data.frame
# options(tibble.print_max = Inf)
Human_GFF
```

```
        seqid  source       type     start       end score strand phase
        <chr>   <chr>      <chr>     <int>     <int> <dbl>  <chr> <dbl>
1  CM000663.2 Genbank     region         1 248956422     0      +     0
2  CM000663.2 Genbank centromere 122026460 125184587     0      +     0
3  KI270706.1 Genbank     region         1    175055     0      +     0
4  KI270707.1 Genbank     region         1     32032     0      +     0
5  KI270708.1 Genbank     region         1    127682     0      +     0
6  KI270709.1 Genbank     region         1     66860     0      +     0
7  KI270710.1 Genbank     region         1     40176     0      +     0
8  KI270711.1 Genbank     region         1     42210     0      +     0
9  KI270712.1 Genbank     region         1    176043     0      +     0
10 KI270713.1 Genbank     region         1     40745     0      +     0
```

### Example `ENSEMBL`:

```{r,eval=FALSE}
# download the GFF file of Homo sapiens from ENSEMBL
# and store the corresponding file in 'ensembl/annotation'
HS.gff.ensembl <- getGFF( db       = "ensembl", 
                         organism = "Homo sapiens", 
                         path = file.path("ensembl","annotation"))
```

After downloading the `.gff` file, users can import the `.gff` file with `read_gff()`.

```{r,eval=FALSE}
# import downloaded GFF file
Human_GFF <- read_gff(file = HS.gff.ensembl)
```

```{r,eval=FALSE}
# show all elements of the data.frame
# options(tibble.print_max = Inf)
Human_GFF
```

```
   seqid source              type start       end   score strand phase
   <int>  <chr>             <chr> <int>     <int>   <chr>  <chr> <dbl>
1      1 GRCh38        chromosome     1 248956422       .      .     0
2      1      . biological_region 10469     11240 1.3e+03      .     0
3      1      . biological_region 10650     10657   0.999      +     0
4      1      . biological_region 10655     10657   0.999      -     0
5      1      . biological_region 10678     10687   0.999      +     0
6      1      . biological_region 10681     10688   0.999      -     0
7      1      . biological_region 10707     10716   0.999      +     0
8      1      . biological_region 10708     10718   0.999      -     0
9      1      . biological_region 10735     10747   0.999      -     0
10     1      . biological_region 10737     10744   0.999      +     0
```

Alternatively for `getGTF()`:

```{r,eval=FALSE}
# download the GTF file of Homo sapiens from ENSEMBL
# and store the corresponding file in 'ensembl/annotation'
HS.gtf.ensembl <- getGTF( db       = "ensembl", 
                         organism = "Homo sapiens", 
                         path = file.path("ensembl","annotation"))
```

### Example `ENSEMBLGENOMES`:

Due to the inavailability of `"Homo sapiens"` at `db = "ensemblgenomes"` here we choose `"Arabidopsis thaliana"`
as example.

```{r,eval=FALSE}
# download the GFF file of Arabidopsis thaliana from ENSEMBLGENOMES
# and store the corresponding file in 'ensemblgenomes/annotation'
AT.gff.ensemblgenomes <- getGFF( db       = "ensemblgenomes", 
                                 organism = "Arabidopsis thaliana", 
                                 path = file.path("ensemblgenomes","annotation"))
```

After downloading the `.gff` file, users can import the `.gff` file with `read_gff()`.

```{r,eval=FALSE}
# import downloaded GFF file
Cress_GFF <- read_gff(file = AT.gff.ensemblgenomes)
```

```{r,eval=FALSE}
# show all elements of the data.frame
options(tibble.print_max = Inf)
Cress_GFF
```

```
   seqid source           type start      end score strand phase
   <int>  <chr>          <chr> <int>    <int> <dbl>  <chr> <dbl>
1      1   TAIR     chromosome     1 30427671     0      .     0
2      1   tair           gene  3631     5899     0      +     0
3      1   tair     transcript  3631     5899     0      +     0
4      1   tair five_prime_UTR  3631     3759     0      +     0
5      1   tair           exon  3631     3913     0      +     0
6      1   tair            CDS  3760     3913     0      +     0
7      1   tair           exon  3996     4276     0      +     0
8      1   tair            CDS  3996     4276     0      +     2
9      1   tair           exon  4486     4605     0      +     0
10     1   tair            CDS  4486     4605     0      +     0
```

Alternatively for `getGTF()`:

```{r,eval=FALSE}
# download the GTF file of Arabidopsis thaliana from ENSEMBLGENOMES
# and store the corresponding file in 'ensemblgenomes/annotation'
AT.gtf.ensemblgenomes <- getGTF( db       = "ensemblgenomes", 
                                 organism = "Arabidopsis thaliana", 
                                 path = file.path("ensemblgenomes","annotation"))
```

Taken together, `getGFF()` or `getGTF()` in combination with `getGenome()`, `getProteome()`, `getRNA()` and `getCDS()` allows users to retrieve the genome information together with the
corresponding `.gff` or `gtf` annotation file to make sure that they both have the same version and origin.

## Repeat Masker Retrieval

In addition to genome, proteome, CDS, etc. information NCBI RefSeq and NCBI Genbank store also 
transposon annotations files generated with Repeat Masker. To retrieve and import
these transposon annotations files, `biomartr` implements the `getRepeatMasker()`
and `read_rm()` functions.

### Example `NCBI RefSeq`:
```{r,eval=FALSE}
# download repeat masker annotation file for A. thaliana
AT.rm.refseq <- getRepeatMasker( db       = "refseq", 
                                 organism = "Arabidopsis thaliana", 
                                 path = file.path("refseq","TEannotation"))
```

Now users can import the TE annotation file using `read_rm()`.

```{r,eval=FALSE}
# import TE annotation file
AT.rm.refseq.anno <- read_rm(AT.rm.refseq)
# look at data
AT.rm.refseq.anno
```

For `NCBI Genbank`, `ENSEMBL`, and `ENSEMBLGENOMES` users can specify the `db` argument analogously as shown above.


## Genome Assembly Stats Retrieval





